Molecular and Functional Analysis of Calcium Binding by a Cancer-linked Calreticulin Mutant

Version 1

Abstract

Calreticulin (CRT) is an endoplasmic reticulum (ER) chaperone with low affinity calcium binding sites in its C-terminal domain. This region is altered by somatic mutations in the CRT gene (CALR), which drive a subset of myeloproliferative neoplasms (MPN). Perturbations in ER calcium storage and signaling are reported for the MPN type I mutant, CRTDel52, and are linked to disease pathogenesis. Using recombinant CRT proteins, we found similar low affinity calcium binding characteristics for wild-type CRT and CRTDel52, as determined by isothermal titration calorimetry (ITC). Residues 340–349, conserved in both wild-type CRT and CRTDel52, contribute to binding. Furthermore, ER and cytosolic calcium levels and store-operated calcium entry (SOCE) were comparable in CRT knockout (CRT-KO) cells reconstituted with wild-type CRT or CRTDel52. Notably, the CRT-KO induces expression of multiple ER proteins known to contain low affinity calcium binding sites or regulate ER calcium levels. Overall, these findings indicate that CRTDel52 retains at least some low affinity calcium binding capacity and that CRT-deficiency induces compensatory cellular changes that maintain ER calcium homeostasis.

Introduction

Calreticulin (CRT) is an endoplasmic reticulum (ER) luminal protein that plays key roles in the assembly and folding of N-linked glycoproteins. CRT has a highly acidic C-terminal domain that contains multiple low affinity calcium binding sites (Baksh and Michalak, 1991, Wijeyesakere et al., 2011). As an ER chaperone, CRT is thought to play a critical role in maintaining cellular calcium homeostasis (Venkatesan et al., 2021b, Michalak, 2024). The ER has a relatively high free calcium concentration of 0.5-1 μM through the presence of various low affinity calcium-binding proteins that buffer calcium, whereas cytosolic calcium levels are maintained within a narrow range of around 100 nM (Michalak, 2024). The maintenance of calcium homeostasis within cellular compartments is essential for cell function and survival (Bagur and Hajnoczky, 2017). Early studies have shown that the C-domain of CRT contributes to ER calcium storage, although free ER luminal concentrations are unchanged (Nakamura et al., 2001).

Somatic mutations of exon 9 of the CALR gene including a 52-base pair deletion (Del52; type 1 mutation) and a 5-base pair insertion (Ins5; type 2 mutation), are driver mutations in some myeloproliferative neoplasms (MPN), specifically in essential thrombocythemia (ET) and primary and secondary myelofibrosis (PMF) (Klampfl et al., 2013, Nangalia et al., 2013). The MPN mutations change the CRT C-domain residues from being predominantly acidic to a high prevalence of basic amino acids and result in a loss of the ER-retention KDEL sequence (Klampfl et al., 2013, Nangalia et al., 2013). Although all MPN CRT mutants change the sequence and charge of the C-domain, a greater number of acidic residues are retained in MPN type 2 mutant (CRTIns5), compared to the type I mutant (CRTDel52). The loss of more acidic residues from the C-terminal domain of CRTDel52 has been reported to impair its calcium binding and reduce ER calcium levels (Ibarra et al., 2022). Store-operated calcium entry (SOCE) is a physiological process that allows for the entry of extracellular calcium and ER calcium restoration following ER calcium depletion (Putney, 2011, Zhang et al., 2020). Altered calcium release from the ER and SOCE was reported for MPN type I but not type 2 mutants in cultured megakaryocytes from MPN patients (Pietra et al., 2016). In another study, spontaneous calcium spikes and increased SOCE activation were observed in cultured megakaryocytes from MPN type I and type 2 patients when compared to healthy donors (Di Buduo et al., 2020).

Although global calcium binding defects were predicted or qualitatively measured for CRTDel52 (Shivarov et al., 2014, Ibarra et al., 2022) and altered cellular calcium signaling is also reported (Pietra et al., 2016, Di Buduo et al., 2020), it remains unclear whether MPN-linked CRTDel52 has any low affinity calcium binding capacity, given that a few conserved acidic residues are retained in its C-domain (Figure 1). In fact, our previous studies showed that a truncated version of murine wild-type CRT (mCRT) containing the non-mutated part of CRTDel52 and two additional C-terminal residues (mCRT1-351) showed similar low affinity calcium binding characteristics as the wild-type mCRT (Wijeyesakere et al., 2011). On the other hand, further truncation of mCRT (mCRT1-339) was needed to disrupt calcium binding (Wijeyesakere et al., 2011, Wijeyesakere et al., 2016). The goal of the current study was to quantitatively measure low affinity calcium binding by CRTDel52 compared with wild-type human CRT and also quantify ER and cytosolic calcium levels and calcium responsiveness in cells expressing wild-type CRT or CRTDel52.

Sequence alignments of the C-terminal residues of murine and human calreticulin (mCRT and hCRT), MPN-linked CRT mutants Ins5 and Del52.

Multiple sequence alignment using Clustal Omega was employed to compare C-domain sequences of mature mCRT, hCRT, Del52, and Ins5 proteins. Acidic residues (340 – 349) previously implicated in low affinity calcium binding by mCRT that are shared between wild-type and both CRTDel52 and CRTIns5 are highlighted in green. Additional acidic residues shared just between wild-type CRT and CRTIns5 are shown in red. There are additional acidic residues unique to wild-type CRT. Mature protein numbering is used for the CRT sequences.

Material and methods

Protein expression and purification

B1 domain of streptococcal protein G (GB1) and his-tagged versions of wild-type CRT, CRT(1- 351), CRT(1-339) and CRTDel52 were purified following expression in E. coli cells using a nickel resin and gel filtration chromatography as described previously (DelProposto et al., 2009, Del Cid et al., 2010, Wijeyesakere et al., 2011) (also see supplemental methods). The monomer fractions were pooled, dialyzed and concentrated to 4 mg/mL in buffer containing 20 mM HEPES, 10 mM NaCl 1 mM CaCl2 pH 7.5 using Amicon ultra concentrators with a molecular weight cut-off of 30,000 Dalton (Millipore) and frozen in 10% glycerol for further use.

Isothermal titration calorimetry

Isothermal titration calorimetry (ITC) measurements were done using the Nano-ITC SV (TA Instruments). Purified CRT monomers were dialyzed into 20 mM HEPES, 10 mM NaCl pH 7.5, brought to a concentration of 100 µM, and then pre-incubated with 50-100 µM CaCl2 in 20 mM HEPES, 10 mM NaCl pH 7.5 for 30 minutes to block the high affinity calcium binding sites. This was followed by 25 sequential 10µL injections of 20 mM HEPES, 10 mM NaCl, 10 mM CaCl2 pH 7.5 into the protein solutions maintained at 37°C. The heat changes measured for the sequential injections were curve-fitted to calculate the enthalpies, stoichiometries, and the dissociation constants using NanoAnalyze software.

Cytosolic calcium measurement and store-operated calcium entry (SOCE) induction

MEG-01 cells and their CALR-modified versions (see supplemental methods) were plated on glass coverslips pretreated with 0.01% poly-L-lysine to facilitate cell attachment. Cells were loaded with 2.5 µM Fura-2/AM dye (Ion Biosciences) supplemented with 0.1% pluronic acid in a physiological salt solution (PSS) containing 140 mM NaCl, 2 mM CaCl2, 5 mM KCl, 2 mM MgCl2, 10 mM HEPES and 11 mM glucose, and incubated for 60 minutes at 37°C. Live single-cell imaging was done at 20X magnification using an Olympus IX71 inverted microscope equipped with a TILL Polychrome V monochromator light source, a Dual-View emission beam splitter, and a cooled Photometrics Quant-EM camera. Cells were transferred to a perfusion chamber that was mounted on the microscope stage, and PSS with or without 2 mM CaCl2 was flowed over cells using a gravity-fed perfusion system. Cytosolic calcium measurements were performed by perfusing Fura-2 loaded cells in the chamber with 2 mM Ca2+, followed by the addition of 25 µM cyclopiazonic acid (CPA) (Cayman) into the imaging buffer. To determine cytosolic calcium levels after activating SOCE, cells were initially perfused with Ca2+-free PSS containing 0.1 mM EGTA, followed by the addition of 25 µM CPA in Ca2+-free PSS and subsequent addition of 2 mM Ca2+ and 25 µM CPA in PSS (without EGTA) to reestablish calcium influx. The emission signals corresponding to the 340 and 380 nm excitation were recorded, and the signals resulting from 340/380 nm excitation were calculated using the Meta Fluor software. The basal cytosolic calcium was defined as the 340/380 nm signal ratio before CPA treatment (measured over a 4-or 5-minute interval immediately before CPA addition). The fractional increase in cytosolic calcium levels after ER Ca2+ depletion with CPA was calculated as the difference between peak signal (single time point) and baseline (averaged over a 4-or 5-minute interval before CPA addition), as a ratio relative to the baseline. The fractional increase in cytosolic calcium after addition of 2 mM extracellular Ca2+ and CPA was calculated as the difference between the peak signal following Ca2+ and CPA addition (averaged over a 1-minute time interval following the peak signal) and the baseline (the signal averaged over a 0.8-1-minute interval immediately before the Ca2+ and CPA addition) relative to the baseline. Total calcium release after ER Ca2+ depletion was reported as the total area under the curve (AUC). Fractional increases were calculated as (Peak-Baseline)/Baseline x 100. This was calculated as peak signals after CPA addition relative to the baseline before CPA (measured over a 4-or 5-minute interval immediately before CPA addition).

ER calcium measurements in HEK293T cells

CALR-modified HEK cells were prepared as described in supplemental methods. Cells were transiently transfected with pCIS-GEM-CEPIA1er (Suzuki et al., 2014); Addgene plasmid # 58217) using Fugene HD reagent (Promega). The transfected cells were harvested after 72 hours. Cells were washed twice with FACS buffer (PBS + 2% FBS) supplemented with 2 mM Ca2+ and resuspended in the same FACS buffer. The live-dead stain 7-AAD (BD) was added at 1:200 dilution and the cells were analyzed by flow cytometry using the BD LSRFortessa cell analyzer. The signal for the Ca2+-bound GEM-CEPIA1er probe was recorded in the Pacific Blue channel (excitation at 405 nm and emission at 452-455 nm) and the signal for the unbound probe was recorded in the AmCyan channel (excitation at 405 and emission at 498 nm). The analysis of flow cytometry data was done in the FlowJo software (v10.10.00). The cells were gated based on the forward (FSC-A) and side scatter (SSC-A), followed by gating on the 7AAD-negative (live) cells, and subsequent gating on the cells positive for the expression of GEM-CEPIA1er probe measured in the Pacific Blue and AmCyan channels. Basal signals were initially recorded, followed by the addition of 5 µM thapsigargin (Tg) and subsequent recording of signals for the treated and untreated cells at four different time points between 2 – 10 minutes. The ratio of mean fluorescence intensity (MFI) of the bound to the unbound signal was used for calculations.

RNA-sequencing

The parental and CRT-KO MEG-01 cells were grown in complete media for 12-18 hours before harvesting. Total RNA was isolated using the RNeasy Plus Mini Kit (Qiagen). This was followed by poly-A selection and enrichment of mRNA using NEB Next Ultra Express RNA kit and a NEB Poly A kit (New England Biolabs). Sequencing was done on the NovaSeq at PE150, targeting 30-40 million reads/sample after cDNA library preparation. Data were pre-filtered to remove genes with 0 counts in all samples. Differential gene expression analysis was performed using DESeq2. The differentially expressed genes were selected as those with a log2 fold-change of <-0.49 or >0.49 and p values <0.05. The results from RNA sequencing were further analyzed in the iPathwayGuide to identify significantly altered proteins related to calcium signaling. The volcano plot was plotted in GraphPad Prism as Log2 fold-change (x-axis) vs. –Log10 adjusted p-value.

Real-time polymerase chain reaction

Total RNA was isolated using the RNeasy Plus Mini Kit (Qiagen) followed by synthesis of complementary DNA using the Superscript III First-Strand Synthesis Kit (ThermoFisher Scientific) following the manufacturer’s protocol. Quantitative RT-PCR was performed on the ABI 7500 FAST Real Time PCR machine (ThermoFisher Scientific, USA) using the SYBR Green PCR Master mix (Applied Biosystems, UK) and gene-specific primers (Supplementary Table 2). Measurement was done in technical duplicates. The expression of target genes was quantified using the 2−ΔΔCt method and normalized to the expression of either one or multiple endogenous control genes including beta-actin (ACTB), glyceraldehyde 3-phosphate dehydrogenase (GAPDH) and hypoxanthine phosphoribosyltransferase 1 (HPRT1). The expression of the target gene in MEG-01 CRT-KO and wild type and CRTDel52 reconstitution cells was further normalized to the mean expression of parental MEG-01 or CRT-KO-vec. The calculated relative mRNA transcript expression values (RQ) were plotted using GraphPad Prism.

Results

MPN-linked CRTDel52 retained at least some low affinity calcium binding sites

Previous studies (Wijeyesakere et al., 2011) have implicated residues 340 to 351 of murine CRT, which are identical in wild-type human CRT and fully conserved with the 340 to 349 residues of CRTDel52, to be involved in low affinity calcium binding. Further truncation of the C-domain to the residue 339 (mCRT1-339) led to the loss of calcium binding measured by ITC (Wijeyesakere et al., 2011). Figure 1 shows a sequence alignment of murine CRT, human CRT, CRTIns5, and CRTDel52 proteins and indicates that the residues previously implicated in low affinity calcium binding are also shared between wild-type CRT and CRTDel52. Recombinant his and GB1 tagged purified proteins were used to compare calcium binding by human CRT and its variants. Chromatograms for human wild-type CRT, CRTDel52, CRT(1-339) and CRT(1-351) are shown in Figure 2A. SDS-PAGE gels corresponding to the purified proteins are shown in Figure 2B. CRTIns5 was not purified in sufficient quantities needed for ITC experiments and was not further investigated in this study.

Purification of recombinant human wild-type CRT, CRTDel52, and C-domain truncation mutants.

A) Representative chromatograms of recombinant GB1-and his-tagged human CRTWT (hCRTWT), CRTDel52, and the C-domain truncation mutants hCRT (1-351) and hCRT (1-339) purified by gel exclusion chromatography after expression in E. coli and initial purification using a nickel resin. B) Representative Coomassie blue-stained SDS-PAGE gels of purified proteins that were used for calcium-binding measurements.

For the ITC measurements, calcium titrations were performed after pre-blocking the high affinity calcium binding sites with 50-100 µM CaCl2. Similar binding profiles and properties were observed with wild-type CRT, CRTDel52, and the CRT(1-351) mutant (Figures 3A, C and D and Table 1). However, consistent with previous studies with mCRT(1–339), endothermic calcium binding by human CRT(1–339) was of lower magnitude compared to other constructs, with a KD value that was too low to be reliably fit, suggesting the loss of low affinity calcium binding sites in this mutant (Figure 3B and Table 1). There are exothermic peaks observed with human CRT(1–339) at the beginning of the calcium titration, which may be attributed to altered low affinity sites retained within the mutant or reduced affinity of the high affinity sites. We concluded that CRTDel52 retains low affinity calcium binding that is measurable by ITC.

Calcium binding to the low affinity sites of purified human CRT proteins.

Low affinity calcium binding by purified hCRTWT, CRTDel52, and the indicated truncation mutants was measured using isothermal titration calorimetry. A-D) Representative isotherm and fitting curves of calcium binding to the low affinity sites of the wild-type and mutant CRT proteins as indicated, measured using isothermal titration calorimetry (ITC). Prior to titration, the proteins were incubated with 50-100 µM CaCl2 to block the high affinity calcium binding sites. At 37°C, 10 mM CaCl2 was titrated against 100 µM protein. The heat change was measured for 25 injections, and the data were curve-fitted to calculate the dissociation constant, enthalpy, and binding stoichiometry using NanoAnalyzer software. Data replicates and statistical analyses are specified in Table 1.

Similar ER calcium levels in CRT-KO HEK cells reconstituted with wild-type CRT or CRTDel52

While the CRTDel52 mutant retains some ability to bind calcium (Figure 3 and Table 1), the loss of the ER-retention KDEL sequence induced by the frameshift mutation, which results in the secretion of the protein from the ER (Araki et al., 2016, Chachoua et al., 2016, Han et al., 2016, Arshad and Cresswell, 2018, Liu et al., 2020, Masubuchi et al., 2020, Venkatesan et al., 2021a, Pecquet et al., 2023) might be expected to cause ER calcium insufficiency in cells expressing CRTDel52 compared to wild-type CRT. We investigated ER calcium signals in CRT knock-out HEK293T cells generated by CRISPR/Cas9 gene editing. A clonal KO line was generated and subsequently reconstituted with constructs encoding wild-type CRT, CRTDel52, or CRTDel52 with a KDEL sequence added at the C-terminus (CRTDel52-KDEL). Protein expression was assessed by immunoblotting with anti-CRT(N), which detects an N-domain epitope present in all CRT constructs, and anti-CRT(C-mut), which was specifically raised against the mutant CRT C-terminus (Venkatesan et al., 2021a). While the recognition with anti-CRT (N) was very weak in lysates of cells expressing CRTDel52, which is secreted, CRTDel52-KDEL is readily detectable with both antibodies. Both wild-type CRT and CRTDel52-KDEL were reconstituted to higher levels than in the parental cells (Figure 4A, lanes 1 and 2 compared to 7-8 and 11-12). The KO and reconstituted cells were further transfected with pCIS-GEM-CEPIA1er, which encodes a ratiometric ER calcium probe (Suzuki et al., 2014). Emission signals for calcium-bound (452-455 nm) and unbound (498 nm) probe were measured using flow cytometry in the Pacific Blue and AmCyan channels, respectively, following excitation with the 405 nm (violet) laser. Flow cytometric analysis was undertaken in a buffer containing 2 mM extracellular Ca2+ (Figures 4B and 4C). There was a significant decrease in basal free ER calcium, measured by the bound/unbound ratios in CRT-KO cells compared to the parental cells, (Figure 4D, left panel). ER calcium levels were partially restored, although non-significantly, when the CRT-KO cells were reconstituted with wild-type CRT. CRT-KO cells reconstituted with CRTDel52 and CRTDel52- KDEL showed small but non-significant decreases in ER calcium levels compared to those reconstituted with wild-type CRT (Figure 4D, right panel). Changes to ER calcium levels in response to thapsigargin (Tg), a SERCA inhibitor, were similar in the parental vs. CRT-KO cells, and in the CRT-KO cells compared with those reconstituted with different CRT proteins (Figure 4E).

ER calcium measurements in CRT-KO HEK293T cells and those reconstituted with wild-type CRT, CRTDel52, or CRTDel52-KDEL.

A) CRISPR-Cas9 editing of HEK cells using lentiviral constructs of calreticulin-targeting gRNA (CRT-KO) or empty vector (HEK-Vec). CRT-KO HEK cells were subsequently reconstituted with wild-type human CRT, CRTDel52, or CRTDel52 with a C-terminal KDEL sequence as indicated. Representative blots showing expression levels of CRT in the indicated HEK cells. Cell lysates were subjected to SDS-PAGE followed by immunoblotting. Wild-type CRT, mutant CRT, and GAPDH were detected with anti-CRT (N), anti-CRT C-(mut), and anti-GAPDH antibodies, respectively. B) Gating strategy for flow cytometry. HEK-Vec cells transiently transfected with the vector encoding the ratiometric GEM-CEPIA1er probe for 72 hours were used to assess ER calcium levels by flow cytometry based on the measurements of fluorescence emission signals for the Ca2+-bound or unbound probe at 452-455 nm and 498 nm, respectively following excitation with the 405 nm (violet) laser. Transfected cells were gated based on forward (FSC-A) and side scatter (SSC-A), then on 7AAD-negative (live) cells, and further on the cells positive for the calcium-bound or unbound probe signals. Representative dot plots of changes in fluorescence intensities in response to treatment with thapsigargin (Tg), a SERCA inhibitor, before (0 mins) or after (10 mins) treatment, are shown. C) Representative histograms showing changes to Ca2+-bound and unbound signals at different time points following Tg treatment of cells. Mean fluorescence intensities (MFI) for Ca2+-bound and unbound signals were used to calculate the bound/unbound ratios at different time points. D and E) The basal free ER calcium signals were measured as bound/unbound MFIs ratios from the corresponding histograms of CRT-KO or HEK-Vec cells (D, left panel) or CRT-KO cells reconstituted with the indicated CRT constructs (D, right panel). The fractional changes in ER calcium signals in response to Tg were measured at different time points in the indicated cells (E). The fractional change ratios were normalized to the baseline and calculated by subtracting the ratio at 10 min from the ratio at 0 mins (R0-R10/R0)x100. Statistical significance was determined using GraphPad Prism and based on paired t-tests (left panels of D and E) or one-way ANOVA analyses with Tukey’s test (right panels of D and E). ns, not significant; *** P value < 0.001. All measurements were undertaken in the presence of 2 mM extracellular Ca2+. Data shown in D and E are based on 7-15 independent experiments as indicated by the individual data points within the graphs. GAPDH, glyceraldehyde-3-phosphate dehydrogenase; SDS-PAGE, sodium dodecyl sulfate-polyacrylamide gel electrophoresis.

Similar cytosolic calcium and SOCE signals in a CRT-KO megakaryocyte cell line reconstituted with wild-type CRT or CRTDel52

MPN CRT mutations cause thrombopoietin-independent differentiation of megakaryocytes to platelets via the thrombopoietin receptor (MPL) (Araki et al., 2016, Chachoua et al., 2016, Marty et al., 2016, Li et al., 2018, Masubuchi et al., 2020). We, therefore, used a megakaryoblast cell line, MEG-01, to assess the role of CRT deficiency and the CRTDel52 mutant on basal cytosolic calcium levels and SERCA inhibition-induced calcium influx into the cytosol. As with the HEK cells (Figure 4), we created a stable CRT knockout clone of MEG-01 cells by CRISPR/Cas9 editing for cytosolic calcium measurements. The cells were subsequently reconstituted with wild-type CRT or CRTDel52. Representative immunoblots are shown to demonstrate the successful knockout and reconstitution (Figure 5A). Wild-type CRT reconstitution in MEG-01 CRT-KO cells resulted in expression levels higher than those in the parental cells (Figure 5A, lanes 7 and 8 compared with lanes 1 and 2). In the presence of 2 mM Ca2+ in the media, we observed no significant differences in the basal cytosolic calcium levels or CPA-induced calcium release from ER in CRT-KO cells compared to the parental MEG-01 cells (Figure 5C and 5D). Additionally, no significant differences in basal cytosolic calcium were observed in CRT-KO cells compared to those reconstituted with wild-type CRT or CRTDel52 (Figure 5E). Furthermore, cytosolic calcium levels in CRT-KO and the two reconstituted cell lines were not statistically different following CPA treatments.

Measurements of cytosolic calcium levels in megakaryoblastic (MEG-01) parental cells, CRT-KO cells, or those reconstituted with wild-type CRT or CRTDel52

Cells in 2 mM extracellular Ca2+ were loaded with the Fura-2/AM dye, and the ratio of the signals was calculated following excitation at 340 and 380 nm and emission at 510 nm. A) CRISPR-Cas9 editing of MEG-01 cells using lentiviral constructs encoding calreticulin-targeting gRNA (CRT-KO) or empty vector (Vec). CRT-KO MEG-01 cells were subsequently reconstituted with viral constructs encoding wild-type human CRT or CRTDel52 or transduced with empty vector (Vec) as indicated in the Figure panels. Representative immunoblots of CRT expression in the indicated cells detected by the anti-CRT(N) and anti-CRT(C-mut) antibodies. B) Representative calcium imaging data from indicated cells in the presence of 2 mM extracellular Ca2+ and following ER calcium depletion with SERCA inhibitor, cyclopiazonic acid (CPA). C and E) Basal cytosolic calcium levels were plotted as the means of ratios of emission signals following excitation at 340 and 380 nm. D and F) Fractional changes to cytosolic calcium levels in the indicated cell lines, measured following treatment with CPA and total calcium release determined as the area under the curve (AUC), are plotted relative to the baseline following CPA addition. Data shown in C-F are based on 4 or 5 independent measurements, each including 50-100 cells. A paired t-test was used to analyze the mean difference between two groups (C and D). Multiple comparisons among groups were analyzed using one-way ANOVA analyses with Tukey’s test (E and F). Statistical significance was determined based p-value ≤ 0.05 using GraphPad Prism. ns, not significant.

SOCE is an important process for restoring ER calcium and maintaining intracellular calcium levels. In contrast to cytosolic calcium signals in the presence of 2 mM Ca2+ in the media, in calcium-free saline, CRT-KO cells exhibited a small but significant increase in basal cytosolic calcium levels relative to the parental cells (Figure 6A, left panel and 6B). Upon ER calcium depletion with CPA, the mean fractional increase in the cytosolic calcium measured in the Ca2+-free saline did not differ significantly in parental cells vs. CRT-KO cells. The areas under the curve were also not different for the two cell lines (Figures 6A and 6C). However, consistent with the higher basal cytosolic calcium measured in CRT-KO cells in calcium-free saline, the measured SOCE signal (following the addition of 2 mM Ca2+ and CPA) was significantly reduced in CRT-KO cells compared to parental cells (Figure 6D). Reconstitution of the CRT-KO cells with either wild-type CRT or CRTDel52 tended to reduce basal cytosolic calcium and induce SOCE (Figure 6A, right panel, 6E and 6G), but both trends were non-significant compared to CRT-KO cells. Importantly, there were no significant differences between the wild-type CRT and CRTDel52 reconstituted cells, in either their basal or CPA-induced cytosolic calcium signals or in SOCE (Figures 6E-G).

Measurements of cytosolic calcium levels and SOCE in MEG-01 parental cells, CRT-KO, or those reconstituted with wild-type CRT or CRTDel52.

Cells were loaded with Fura-2 dye, and the ratio of the 340 nm/380 nm signals was calculated following excitation at 340 and 380 nm and emission at 510 nm. A) Representative calcium imaging data from cells in the absence of extracellular calcium, followed by the addition of CPA, and subsequent addition of extracellular calcium with CPA as indicated. B and E) Compiled basal cytosolic calcium levels in the indicated cells in the absence of extracellular calcium plotted as the ratios of 340 nm/380 nm signals. C and F) Fractional changes to cytosolic calcium and total calcium release measured as the area under the curve (AUC) in the indicated cells following ER calcium depletion with the CPA. D and G) Fractional changes to cytosolic calcium following ER calcium depletion with the CPA and subsequent addition of extracellular calcium with CPA. Data are plotted as a fractional increase relative to baseline following CPA addition. SOCE in the indicated cells is plotted as the fractional increase in signal following the addition of 2 mM extracellular calcium in the presence of CPA. Data shown are based on 8-10 independent measurements, each with 20-100 cells. Comparisons were performed by paired t-test and one-way ANOVA analysis using GraphPad Prism. Statistical significance is based on a p-value ≤ 0.05. ns, not significant.

CRT deficiency induced numerous transcriptional changes to genes within the calcium signaling pathway

As noted in Figures 4 and 6, some of the measured differences between CRT-KO and parental cells are not fully reversed by reconstitution with wild-type CRT, although the trends are maintained. We predicted that there may be other indirect effects of CRT-KO and reconstitution that affect cellular calcium levels and calcium signaling that could explain some of these measured differences. As proof of concept, we undertook bulk RNA-sequencing to determine differential gene expression in MEG-01 parental vs. CRT-KO cells. There were 3078 upregulated and 1790 downregulated genes. The volcano plot and heat map show identified genes that were differentially expressed in CRT-KO MEG-01 compared to parental cells (Figure 7A). Based on pathway analysis, numerous genes involved in cellular calcium signaling were identified. Of specific note is the upregulation of GRP94 encoded by the HSP90B1 gene, BiP encoded by HSPA5 gene, and PDIA6 encoded by PDIA6 gene, which are known to have low affinity calcium binding sites (Lievremont et al., 1997, Macer and Koch, 1988, Biswas et al., 2007, Marzec et al., 2012, Okumura et al., 2021) (Figures 7A and 7B). Other studies have reported similar upregulation of these genes in CRT-KO (Tang et al., 2025) as well as in heterozygous CRTDel52 knock-in cells (Figure 7C) (Fosselteder et al., 2023). Upregulation of these proteins may serve a compensatory role in buffering calcium in CRT-KO cells. Additionally, PDIA3, which is shown to regulate the activities of stromal interaction molecule 1 (STIM1) (Prins et al., 2011) and Sarco/endoplasmic reticulum Ca2+ ATPase (SERCA2b) (Li and Camacho, 2004), was consistently upregulated in all the studies.

Transcriptional changes to the expression of genes encoding ER calcium binding/regulating proteins in CRT-KO MEG-01 cells.

A) RNA sequencing analysis was used to identify differentially expressed genes in Meg-01 CRT-KO cells in comparison to parental Meg-01 cells. Volcano plot shows significantly upregulated (blue) and downregulated genes (red) genes identified by RNA sequencing. Genes related to ER calcium binding and homeostasis have been highlighted and labeled. The differentially expressed genes were selected as those with a log2 fold-change of <-0.49 or >0.49 and p values <0.05. B) Heat map shows the differential expression of genes encoding ER resident calcium binding proteins or ER calcium regulators that were identified by RNA sequencing in CRT-KO Meg-01 cells compared to parental cells (n=3 biological replicates). C) Comparison of expression ratio of HSPA5, HSP90B1, PDIA3 and PDIA6 in CRT-KO Meg-01 cells (this study) or murine lung cancer cells (Tang et al., 2025) or heterozygous CRTDel52 knock-in (Del52-KI) cells (Fosselteder et al., 2023) normalized to their expression in respective control cell lines expressing wild type CRT. The normalized expression of various genes identified in each dataset was calculated as a ratio of expression in CRT-KO or CRTDel52 knock-in cells relative to the expression in wild-type CRT-expressing cells. Multiple unpaired t-tests were used to determine statistical significance with the Benjamini-Hochberg FDR correction (p-adjusted) <0.05. D-G) Left and middle panels: Relative expression of mRNA transcripts of HSPA5 (n=8-11), HSP90B1 (n=7), PDIA6 (n=6) and PDIA3 (n=6) genes in the indicated MEG-01 cells determined by RT-qPCR using specific gene primers. Relative gene expression was calculated by normalizing the target gene’s cycle threshold (Ct) to a single or multiple endogenous controls (ACTB, GAPDH and HPRT1) which was used for comparison across different cell lines as indicated. Data show 6 – 11 independent runs with duplicate measurements for each run. Student t-test was used to determine the statistical significance between two groups while multiple group comparison was performed by one-way ANOVA analysis using GraphPad Prism. Statistical significance is based on a p-value ≤ 0.05. ns, not significant. Right panels: Representative immunoblots of BiP, GRP94, PDIA6 and PDIA3 expression in the indicated cells detected by specific antibodies.

We further used real-time PCR with target gene primers (Supplementary Table 2) to validate the expression of the key ER resident proteins identified from the RNA-sequencing data which have roles in calcium binding and regulation of ER calcium homeostasis. Consistent with the RNA-seq data, the expression of HSPA5, HSP90B1, PDIA6 and PDIA3 mRNA transcripts was significantly increased in CRT-KO MEG-01 compared to parental cells. Reconstitution of CRT-KO cells with wild-type CRT but not CRTDel52 showed a significant decrease in the expression of HSPA5, HSP90B1, PDIA6 and PDIA3 (Figures 7D-G). Protein levels of BiP, GRP94, PDIA6 and PDIA3 were also assessed in the indicated cells by immunoblots using specific antibodies (supplementary methods). Representative immunoblots show increased expression of BiP (Figure 7D, right panel), GRP94 (Figure 7E, right panel), PDIA6 (Figure 7F, right panel) and PDIA3 (Figure 7G, right panel) in CRT-KO cells compared to parental MEG-01 cells as well as increased expression in CRT-KO-Vec cells compared to CRT-KO cells reconstituted with wild type CRT. The effects of CRTDel52 in reducing expression were generally less robust than wild type CRT.

Discussion

Calreticulin is a ubiquitously expressed ER luminal calcium-binding chaperone, with a role in ER calcium storage (Venkatesan et al., 2021b, Michalak, 2024). Previous studies using dye-based assays reported impaired calcium binding by the MPN-associated CRTDel52 mutant (Ibarra et al., 2022). Yet direct and quantitative analyses of its calcium-binding properties have been lacking. In this study, we investigated low affinity calcium binding by wild-type human CRT, CRTDel52, and the truncation mutants CRT(1-351) and CRT(1-339) using ITC (Figure 3 and Table 1). Our studies show that wild-type CRT, CRTDel52, and CRT(1-351) retain low affinity calcium binding sites, exhibiting average KD values of 574.24, 938.82, and 444.34 µM, respectively. On the other hand, low affinity calcium binding by CRT(1-339) could not be reliably fit. This is the first quantitative characterization of the low affinity calcium binding profiles of wild-type human CRT, CRTDel52, and functionally informative truncation mutants.

Although CRTDel52 retains at least partial low affinity calcium binding capability, loss of its KDEL retention sequence (Figure 1) could still disrupt ER calcium storage and signaling. Interestingly, the cellular assays reveal that ER and cytosolic calcium levels, as well as SOCE, are similar between CRT-KO cells and those reconstituted with either wild-type CRT or CRTDel52 (Figures 4-6). One study has reported decreased ER calcium levels in a human osteosarcoma cell line expressing CRTDel52 compared to those expressing wild-type CRT, following transient transfections (Ibarra et al., 2022). The same study suggested the contributions of the P and C-domain of wild-type in restoring ER calcium levels in cells expressing CRTDel52 (Ibarra et al., 2022). While the conditions of those experiments differ from the studies described here which use a CRT-KO background, our findings are consistent with the prediction that any excess free ER calcium generated in cells lacking CRT or sufficient ER calcium binding sites would rapidly diffuse out of the ER and be pumped out of the cell.

Related to cytosolic calcium levels and SOCE activation, consistent with our findings (Figure 6G), a previous study found no differences in cytosolic calcium levels or SOCE induction in myeloid progenitor 32D cells expressing wild-type CRT and CRTDel52 both under conditions of cytokine deprivation and thrombopoietin (TPO) stimulation (Bhuria et al., 2024). While cultured megakaryocytes from MPN patients with CALR mutations have SOCE abnormalities compared to healthy control cells (Pietra et al., 2016, Di Buduo et al., 2020) sample to sample heterogeneities in primary cells make it difficult to attribute measured differences solely to calcium binding by CRT or its mutants.

We found that CRT-KO HEK cells displayed lower ER calcium compared with the parental HEK cells (Figure 4D, left panel) and also that CRT-KO Meg-01 cells displayed increased cytosolic calcium in the absence of extracellular calcium (Figure 6B). However, reconstitutions with wild-type CRT did not reverse these effects (Figure 4D, right panel and Figure 6E). Transcriptomic analysis identified significant alterations in the expression of genes related to calcium storage and signaling between parental and CRT-KO Meg-01 cells (Figure 7). These changes highlight adaptive or compensatory pathways induced by CRT-KO and emphasize that, when comparing wild-type and CRT-KO cell lines for subcellular calcium levels or responses to SERCA inhibitors, the measured effects might not be based solely on the abilities of CRT to maintain ER and cytosolic Ca2+ levels via its calcium binding, storage and buffering capacity. Rather, many other compensatory cellular changes could affect subcellular calcium levels and signaling. In this study, various ER proteins with known calcium binding activities were found to be upregulated in CRT-KO cells (Figure 7A and 7B). GRP94 (Macer and Koch, 1988, Biswas et al., 2007, Marzec et al., 2012), BiP ((Lievremont et al., 1997) and our unpublished findings), and PDIA6 (Okumura et al., 2021) are each known to contain low affinity calcium binding sites. CRT-KO cells have significantly induced expression of the genes encoding these proteins. Wild type CRT reconstitution into the CRT-KO cells significantly reduced the expression of these ER proteins, whereas CRTDel52 reconstitution only partly and non-significantly reduced expression of HSP90B1 and PDIA6 and did not reduce the expression of HSPA5 (Figures 7D-F). GRP94, BiP and PDIA6 have functions in ER protein folding and quality control (Hendershot et al., 2024, Okumura et al., 2021). It is known that CRTDel52 has impaired chaperone activity towards its substrates compared with wild type CRT (Arshad and Cresswell, 2018, Schurch et al., 2022, Kaur et al., 2025). This could account for some of these measured differences between wild type CRT and CRTDel52 in altering the expression of the ER chaperones and folding factors. However, their induction in knock-in cells with heterozygous expression of wild type CRT and CRTDel52 (Fosselteder et al., 2023), in which the expressed wild type CRT appears sufficient for multiple protein folding activities (Kaur et al., 2025, Schurch et al., 2022) could point to compensatory effects relevant to calcium signaling. Regardless of the precise mechanisms relevant to induction of the calcium binding proteins, their combined expression, despite protein compositional differences between the KO and reconstituted cells, appears sufficient to maintain ER calcium levels (Figure 4D, right panel), calcium release from the ER (Figure 4E, right panel, and 5F) and cytosolic calcium levels (Figure 5E).

PDIA3 (ERp57) forms a complex with CRT and interacts with and regulates the activities of the ubiquitously expressed SERCA2b (Li and Camacho, 2004). Its expression is also induced by the CRT-KO and suppressed by wild type CRT expression but non-significantly by CRTDel52 expression. Altered SERCA2b activity could also affect basal ER calcium levels, contributing to similarities between CRT-KO, and wild type CRT or CRTDel52 reconstituted cells.

Collectively, our findings reveal that CRTDel52 retains some ability to bind calcium at low affinity sites. CRT-KO cells reconstituted with wild-type or CRTDel52 have comparable ER and cytosolic calcium levels and store-operated calcium entry (SOCE). CRT-KO induces broad changes in ER calcium binding protein expression as well as in the expression of ERp57. Overall, we can conclude that a number of factors that contribute to the maintenance of ER calcium levels and regulation of calcium transport into the ER are altered in CRT-KO cells. Thus, any disease-related changes in calcium signaling and homeostasis are likely multifactorial, encompassing transcriptional regulation and altered expression of critical mediators of calcium storage, transport and signaling pathways. These studies provide a more nuanced understanding of CRT’s role in cellular calcium signaling and may inform future therapeutic strategies targeting calcium signaling pathways in hematological malignancies.

eLife Assessment

This study investigates low-affinity Ca2+ binding by WT calreticulin and mutant calreticulin associated with type I myeloproliferative neoplasms, as well as the impact on Ca2+ fluxes in suspension cultures of megakaryocyte-like cells in vitro in response to ER Ca2+ ATPase inhibitors that deplete endoplasmic reticulum (ER) Ca2+ store and open plasma membrane Ca2+ channels through STIM1-Orai interactions. The results are important in that they show that Ca2+ binding by calreticulin and store-operated Ca2+ entry are not fundamentally impacted by the type I deletion mutation in calreticulin, which rules out a direct effect of the calreticulin mutation on its own low-affinity Ca2+ binding and any broad impact on ER Ca2+ regulation. The strength of the data and methods used ranges from solid to convincing, although the use of suspension-based flow cytometric assays to investigate ER Ca2+ levels and Ca2+ entry can be challenged. High-affinity Ca2+ binding sites could be further considered, and possible confounding effects of Abl kinase activity in the megakaryocyte-like cell lines could be offset.

Reviewer #1 (Public review):

The authors attempted to compare calcium calcium-binding properties of wildtype calreticulin with calreticulin deletion mutant (CRTDel52) associated with myeloproliferative neoplasms.

The researchers conducted their study using advanced techniques. They found almost no difference in calcium binding between the two proteins and observed no impact on calcium signaling, specifically store-operated calcium entry (SOCE). The study also noted an increase in ER luminal calcium-binding chaperone proteins. Surprisingly, the authors selected flow cytometry as a technique for measurements of ER luminal calcium. Considering the limitations of this approach, it would be better to use alternative approaches. This is particularly important as previous reports, using cells from MPN patients, indicate reduced ER luminal calcium and effects on SOCE (Blood, 2020). This issue matters because earlier research with MPN patient cells reported reduced ER luminal calcium levels and altered SOCE (Blood, 2020). How do the authors explain the difference between their results and previous findings about lower ER luminal calcium and changed SOCE in MPN patient cells expressing CRTDel52? Other studies have found that unfolded protein responses are activated in MPN cells with CRTDel52 calreticulin (see Blood, 2021), and increased UPR could account for higher levels of some ER-resident calcium-binding proteins observed here. Overall, it remains unclear how this work improves our understanding of MPN or clarifies calreticulin's role in MPN pathophysiology.

Reviewer #2 (Public review):

Summary:

Tagoe and colleagues present a thorough analysis of the calcium (Ca2+) binding capacity of calreticulin (CRT), an endoplasmic reticulum (ER) Ca2+-buffer protein, using a mutant version (CRT del52) found in myeloproliferative neoplasms (MPNs). The authors use purified human CRT protein variants, CRT-KO cell lines, and an MPN cell line to elucidate the differing Ca2+ dynamics, both on the level of the protein and on cell-wide Ca2+-governed processes. In sum, the authors provide new insights into CRT that can be applied to both normal and malignant cell biology.

First, the authors purify CRT protein and perform isothermal titration calorimetry to quantify the Ca2+ binding capacity of CRT. They use full-length human CRT, CRT del52, and two truncations of CRT (1-339 and 1-351, the former of which should lead to the entire loss of low-affinity Ca2+ binding). While CRT del52 has previously been shown to lead to a decrease in Ca2+ binding affinity in other models, the ITC data show that this is retained in CRT del52.

Next, the authors utilize a CRT-KO cell line with subsequent addition of CRT protein variants to validate these findings with flow cytometric analysis. Cells were transfected with a ratiometric ER Ca2+ probe, and fluorescence indicates that CRT del52 is unable to restore basal ER Ca2+ levels to the same extent as CRT wild-type. To translate these findings to MPNs, the authors perform CRT-KO in a megakaryocytic cell line, where reconstitution with either CRT variant did not cause a difference in cytosolic calcium levels. The authors further test store-operated calcium entry (SOCE), an important process for maintaining ER Ca2+ levels, in these cells, and find that CRT-KO cells have lower SOCE activity, and that this can be slightly recovered with CRT addition.

Finally, the authors ask whether other effects of CRT-KO/reconstitution can affect the cellular Ca2+ signaling pathway and levels. RNASeq analysis revealed that CRT-KO leads to an increase in various chaperone protein expressions, and that reconstitution with CRT del52 is unable to reduce expression to the same extent as reconstitution with CRT wildtype.

Strengths:

The authors provide new insights into CRT that can be applied to both normal and malignant cell biology.

Weaknesses:

(1) The authors should consider discussing the high-affinity Ca2+ binding site more in the introduction. Can they show a proof-of-concept experiment that validates that incubation of recombinant CRT reduces the function of that high-affinity Ca2+ binding site?

(2) For Figure 2B, do you have an explanation for why the purified proteins run higher than predicted (48-52kDa) - are these proteins still tagged with pGB1?

(3) The MEG-01 cell line has the BCR::ABL1 translocation, while CRT mutations are strictly found in BCR::ABL1 negative MPNs. Could these experiments be repeated in these cells treated with imatinib to decrease these effects, or see if basal MEG-01 Ca2+ levels/activity are changed with or without imatinib?

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This study investigates low-affinity Ca2+ binding by WT calreticulin and mutant calreticulin associated with type I myeloproliferative neoplasms, as well as the impact on Ca2+ fluxes in suspension cultures of megakaryocyte-like cells in vitro in response to ER Ca2+ ATPase inhibitors that deplete endoplasmic reticulum (ER) Ca2+ store and open plasma membrane Ca2+ channels through STIM1-Orai interactions. The results are important in that they show that Ca2+ binding by calreticulin and store-operated Ca2+ entry are not fundamentally impacted by the type I deletion mutation in calreticulin, which rules out a direct effect of the calreticulin mutation on its own low-affinity Ca2+ binding and any broad impact on ER Ca2+ regulation. The strength of the data and methods used ranges from solid to convincing, although the use of suspension-based flow cytometric assays to investigate ER Ca2+ levels and Ca2+ entry can be challenged. High-affinity Ca2+ binding sites could be further considered, and possible confounding effects of Abl kinase activity in the megakaryocyte-like cell lines could be offset.

The authors thank the editors and the reviewers for the summary, comments and many helpful suggestions. In the revised manuscript, we have used fluorimetry for more precise ER calcium measurements (new Figure 5), clarified the high-affinity calcium binding site concern based on our previous work, and addressed possible effects of BCR-ABL translocation kinase activity upon cellular calcium signaling using the drug Imatinib (new supplements to Figure 6 and 7).

Public Reviews:

Reviewer #1 (Public review):

The researchers conducted their study using advanced techniques. They found almost no difference in calcium binding between the two proteins and observed no impact on calcium signaling, specifically store-operated calcium entry (SOCE). The study also noted an increase in ER luminal calcium-binding chaperone proteins. Surprisingly, the authors selected flow cytometry as a technique for measurements of ER luminal calcium. Considering the limitations of this approach, it would be better to use alternative approaches.

Thank you for this suggestion. We have undertaken fluorimetry-based ER calcium measurements, which are shown in a new Figure 5. These also indicate similar ER calcium levels in CRT-KO HEK293T cells, compared to those reconstituted with wild-type CRT and CRTDel52.

This is particularly important as previous reports, using cells from MPN patients, indicate reduced ER luminal calcium and effects on SOCE (Blood, 2020). This issue matters because earlier research with MPN patient cells reported reduced ER luminal calcium levels and altered SOCE (Blood, 2020). How do the authors explain the difference between their results and previous findings about lower ER luminal calcium and changed SOCE in MPN patient cells expressing CRTDel52?

We thank the reviewer for asking for these clarifications. We have revised the discussion to address some of these points and also clarify the findings of the referenced study (Di Buduo et al., 2020) which did not directly measure ER calcium levels. We also discuss findings from a related study with cultured megakaryocytes from patients that indicated different effects of type I vs type II mutations (Pietra et al., 2016). In the absence of engineered controls, sample-to-sample heterogeneities in primary cells make it difficult to attribute any measured differences as direct effects of CRT mutations. Different from these experiments, by using purified proteins and ITC, our studies show that the Del52 mutant has calcium-binding characteristics resembling that of the wild-type protein. Additionally, through genetic manipulations in cell lines, our studies directly address the effects of calreticulin KO and its Del52 mutation upon ER luminal and cytosolic calcium levels, and cellular SOCE signals. We did not measure significant differences in any of these parameters between the KO cells and those reconstituted with wild-type calreticulin or the Del52 mutant. As noted by the editors, these results show that Ca2+ binding by calreticulin and SOCE in a cell are not fundamentally impacted by the type I deletion mutation.

Other studies have found that unfolded protein responses are activated in MPN cells with CRTDel52 calreticulin (see Blood, 2021), and increased UPR could account for higher levels of some ER-resident calcium-binding proteins observed here.

These points are addressed in the discussion. Either protein misfolding in cells with wild-type calreticulin deficiency or the sensing of cellular calcium perturbations could induce the expression of ER calcium-binding proteins in calreticulin-deficient cells, although we favor the latter model for the reason specified in the discussion. Regardless of the precise mechanisms underlying the expression changes in calcium-binding proteins, the upregulated factors are predicted to compensate for calreticulin deficiency and contribute to the maintenance of the overall cellular calcium homeostasis.

Overall, it remains unclear how this work improves our understanding of MPN or clarifies calreticulin's role in MPN pathophysiology.

Multiple studies referenced in the manuscript have suggested links between altered calcium signaling/binding by CRT mutants and MPN pathogenesis. Our studies indicate that ER and cytosolic calcium levels and SOCE are not directly impacted by the MPN type I CALR mutation, points noted in the abstract and discussion. Thus, calcium signaling may not play a specific role in MPN CALR mutant pathology via suggested mechanisms. We are confident that readers will find these results important for better understanding the role of calreticulin type I mutations in MPN.

Reviewer #1 (Recommendations for the authors):

This study aimed to express, purify, and evaluate low-affinity calcium binding by a calreticulin deletion mutant (CRTdel52) that is linked to myeloproliferative neoplasms (MPN). The researchers performed cell imaging, flow cytometry, and isothermal titration calorimetry to compare calcium binding between wild-type calreticulin and CRTDel52. They assessed cytosolic calcium levels and store-operated calcium entry (SOCE) in HET293T cells (CRT knocked-out background) and in megakaryoblastic MEG-1 cells. Additionally, they examined changes in the abundance of endoplasmic reticulum (ER) resident calcium-binding proteins in cells expressing either wild-type or mutant CRT.

This study is well executed but lacks clear relevance to MPN, and it is not clear how this work advances our knowledge of calreticulin biology. No differences were found in calcium binding between wild-type calreticulin and CRTDel52, nor was SOCE impacted. They noticed, however, a compensatory increase in the abundance of some ER resident calcium-binding proteins. The lack of any significant changes in calcium behavior between wild type and CRTDel52 is not surprising based on the known amino acid sequence of calreticulin and calreticulin mutant and based on our knowledge about CRT calcium binding in general. Consequently, it is not clear how this work advances our understanding of the pathophysiology of MPN. Calcium may not play a critical role in the MPN pathology; instead, CRTDel52 secretion and receptor signaling appear more central. Further research should address how these findings relate specifically to MPN and cell biology, in general.

Our current studies demonstrate increased expression of other calcium-binding proteins in the context of heterozygous MPN type I CALR mutations (Figure 8C) or conditions resembling homozygous MPN type I CALR mutations (Figures 8D-8G). These results, together with findings of maintained ER and cytosolic calcium levels and SOCE signals (Figures 4-7 and Figure 6, supplemental Figure 2 and Figure 7, supplemental Figure 1), indicate that altered calcium binding/signaling by Del52 does not directly contribute to MPN pathology.

What is the biological or pathophysiological relevance of the CRTDel52-KDEL construct?

The KDEL sequence is important for the ER retention of CRT (Sonnichsen et al., 1994), and its addition was expected to at least partially remedy the ER retention defect of CRTDel52. This point is clarified in the revised results section.

The rise in ER calcium-binding proteins is noteworthy but anticipated, given likely genetic changes from UPR pathway activation in these cells. Is this relevant to MPN?

We suggest that increased expression of other calcium-binding proteins in the context of heterozygous MPN type I CALR mutations (Figure 8C) or conditions resembling homozygous MPN type I CALR mutations (Figures 8D-8G) would contribute to the maintenance of the cell’s calcium signaling capacity.

How do the authors explain the difference between their results and previous findings about lower ER luminal calcium and changed SOCE in MPN patient cells expressing CRTDel52?

The findings related to SOCE are addressed in the points discussed above and in the revised discussion. Related to ER luminal calcium, the study by Ibarra et al. (Ibarra et al., 2022) reported that CRTDel52 overexpressed in U2OS cells (expressing endogenous CRT) had reduced ER calcium levels compared to the same cells expressing WT CRT or CRTIns5. Those measurements did not use a ratiometric ER calcium probe, and additionally it is possible that the expression of compensatory calcium-binding proteins is more muted in cells expressing endogenous wild-type CRT.

Other studies have found that unfolded protein responses are activated in MPN cells with CRTDel52 calreticulin (see Blood, 2021), and increased UPR could account for higher levels of some ER-resident calcium-binding proteins observed here.

We agree that increased UPR could account for higher levels of some ER-resident calcium-binding proteins. As noted in the revised discussion, regardless of the precise mechanisms underlying the expression changes in calcium-binding proteins, the upregulated factors are predicted to compensate for calreticulin deficiency and contribute to the maintenance of the overall cellular calcium homeostasis.

It is not clear why flow cytometry was a choice of technique for measurements of ER calcium. Pacific Blue was detected at 405 nm excitation and 452-455 nm emission, while unbound probe signals appeared in the AmCyan channel (405 nm excitation/498 nm emission). It is unnecessary to mention fluorochrome labels (like Pacific Blue or AmCyan) for channels that are not in use. Only the channels actually utilized need to be specified: The calcium-bound GEM-CEPIA1er probe's signal was collected using the Pacific Blue channel (excitation at 405 nm, emission at 452-455 nm), while the unbound probe's signal was detected using the AmCyan channel (excitation at 405 nm, emission at 498 nm). Since it's not possible to monitor emission only at a specific wavelength with a Fortessa, could this be a different channel that is being recorded?

Additionally, due to the similar spectra and potential for bleed-through between Pacific blue and Amcyan, compensation is likely necessary. Therefore, you should include single-stained control cells containing only one probe for proper reporting. Additionally, only the "GEM-CEPIA1er probe" is displayed, while the second probe is referred to solely as "unbound".

A single genetically encoded GEM-CEPIA1er probe (Suzuki et al., 2014) was used for measuring both the bound and unbound signals. The GEM-CEPIA1er probe was excited with the 405 nm violet laser. In the methods section of the revised manuscript, the GEM-CEPIA1er probe wording is included for describing both the bound and unbound signal collections. Additionally, we have undertaken new spectrofluorimetric experiments (new Figure 5), which allow for the distinct emission peaks to be recorded corresponding to the Ca2+-bound and Ca2+-unbound signals. Similar results were obtained as reported for the flow cytometry-based experiments.

Increased expression of wild-type calreticulin compared to parental cells should impact on ER calcium content and dynamics in back-transfected HEK293T-KO or MEG-1 cells. Direct ER calcium measurements in HEK293 cells with various calreticulin constructs would significantly strengthen this presentation.

Our experiments were structured to compare calcium signaling in cells expressing only wild-type CRT or CRTDel52 (resembling homozygous type I MPN CALR mutations) compared to CRT-KO cells. The over-expression of CRT in the reconstituted cells compared to endogenous expression level is a limitation of our study which we have acknowledged in the revised manuscript discussion. Understanding the effects of over-expression of wild-type CRT vs the CRTDel52 mutant upon ER and cytosolic calcium signals and SOCE is interesting, but beyond the scope of the present study.

The authors should examine the immunolocalization of CRTDel52 and wild-type protein in HEK293 cells.

Previous published studies from another lab showed that CRTDel52 is secreted from HEK cells and that the addition of a KDEL sequence to CRTDel52 reduces secretion and induces its increased intracellular accumulation (Arshad and Cresswell, 2018). This point is noted in the revised results section and the reference is cited. This appears to be the general theme in primary cells and cell lines. Previous studies and our own prior published studies have shown that CRTDel52 (but not wild-type CRT) is detectable in the media of cell lines and patient serum as well on the cell surface of primary cells and cell lines (Kaur et al., 2024, Venkatesan et al., 2021, Pecquet et al., 2023).

SDS-PAGE of purified proteins is overloaded, and chromatograms show extra peaks or shoulders, making protein quality assessment uncertain.

Representative chromatograms, peaks corresponding to protein monomers used for ITC analyses and the relevant gels are clarified in the revised manuscript. In new analyses since the original submission, intact protein mass spectrometry was undertaken for CRTDel52. The results indicate a 35-42 amino acid truncation in different preparations. The truncated proteins would still include acidic residues (between 340–351) previously implicated in low-affinity calcium binding by murine CRT that are shared between wild-type and CRTDel52. This new information is now included in the revised results section.

Analysis of SOCE in calreticulin-deficient cells and cells reconstituted with calreticulin or overexpressing the protein has already been reported (PMID12324449).

The indicated reference and additional related papers examining effects of CRT deficiency and overexpression on cellular calcium signaling (Arnaudeau et al., 2002, Bastianutto et al., 1995, Mery et al., 1996, Nakamura et al., 2001) are cited in the revised manuscript.

Reviewer #2 (Public review):

Tagoe and colleagues present a thorough analysis of the calcium (Ca2+) binding capacity of calreticulin (CRT), an endoplasmic reticulum (ER) Ca2+-buffer protein, using a mutant version (CRT del52) found in myeloproliferative neoplasms (MPNs). The authors use purified human CRT protein variants, CRT-KO cell lines, and an MPN cell line to elucidate the differing Ca2+ dynamics, both on the level of the protein and on cell-wide Ca2+-governed processes. In sum, the authors provide new insights into CRT that can be applied to both normal and malignant cell biology.

First, the authors purify CRT protein and perform isothermal titration calorimetry to quantify the Ca2+ binding capacity of CRT. They use full-length human CRT, CRT del52, and two truncations of CRT (1-339 and 1-351, the former of which should lead to the entire loss of low-affinity Ca2+ binding). While CRT del52 has previously been shown to lead to a decrease in Ca2+ binding affinity in other models, the ITC data show that this is retained in CRT del52.

Next, the authors utilize a CRT-KO cell line with subsequent addition of CRT protein variants to validate these findings with flow cytometric analysis. Cells were transfected with a ratiometric ER Ca2+ probe, and fluorescence indicates that CRT del52 is unable to restore basal ER Ca2+ levels to the same extent as CRT wild-type. To translate these findings to MPNs, the authors perform CRT-KO in a megakaryocytic cell line, where reconstitution with either CRT variant did not cause a difference in cytosolic calcium levels. The authors further test store-operated calcium entry (SOCE), an important process for maintaining ER Ca2+ levels, in these cells, and find that CRT-KO cells have lower SOCE activity, and that this can be slightly recovered with CRT addition.

Finally, the authors ask whether other effects of CRT-KO/reconstitution can affect the cellular Ca2+ signaling pathway and levels. RNASeq analysis revealed that CRT-KO leads to an increase in various chaperone protein expressions, and that reconstitution with CRT del52 is unable to reduce expression to the same extent as reconstitution with CRT wildtype.

Strengths:

The authors provide new insights into CRT that can be applied to both normal and malignant cell biology.

We thank the reviewer for the recognition that this study is important for our understanding of both normal and malignant cell biology.

Weaknesses:

(1) The authors should consider discussing the high-affinity Ca2+ binding site more in the introduction. Can they show a proof-of-concept experiment that validates that incubation of recombinant CRT reduces the function of that high-affinity Ca2+ binding site?

In a previous study (Wijeyesakere et al., 2011), we showed that at a starting calcium concentration of 0 mM and with CaCl2 injections to a final concentration of 70-80 mM the measured KD value was 16.6 mM for calcium binding to wild type murine calreticulin, (which has ~95% sequence identity with human calreticulin), corresponding to the high-affinity site. On the other hand, at a starting calcium concentration of 50-100 mM and CaCl2 injections to a final concentration of 700-850 mM, the measured KD value for calcium binding to wild-type murine calreticulin was 590 mM (corresponding to the low-affinity sites). We did not observe the high-affinity sites when the starting calcium concentration was 50 mM and calcium injections were at 33 mM each; similar conditions are used in the present study. These points are clarified in the revised manuscript in the results section.

(2) For Figure 2B, do you have an explanation for why the purified proteins run higher than predicted (48-52kDa) - are these proteins still tagged with pGB1?

Yes, the purified proteins shown in Figure 2B retained a GB1 tag. This point is clarified in the revised methods.

(3) The MEG-01 cell line has the BCR:ABL1 translocation, while CRT mutations are strictly found in BCR:ABL1 negative MPNs. Could these experiments be repeated in these cells treated with imatinib to decrease these effects, or see if basal MEG-01 Ca2+ levels/activity are changed with or without imatinib?

Thank you for this important point. We have assessed cytosolic calcium levels in MEG-01 cells that were treated or not treated with imatinib in new Figure 6, supplemental Figures 1 and 2 and Figure 7, supplemental Figure 1) and show that the prior results hold in imatinib-treated cells.

References

ARNAUDEAU, S., FRIEDEN, M., NAKAMURA, K., CASTELBOU, C., MICHALAK, M. & DEMAUREX, N. 2002. Calreticulin differentially modulates calcium uptake and release in the endoplasmic reticulum and mitochondria. J Biol Chem, 277, 46696-705.

ARSHAD, N. & CRESSWELL, P. 2018. Tumor-associated calreticulin variants functionally compromise the peptide loading complex and impair its recruitment of MHC-I. J Biol Chem, 293, 9555-9569.

BASTIANUTTO, C., CLEMENTI, E., CODAZZI, F., PODINI, P., DE GIORGI, F., RIZZUTO, R., MELDOLESI, J. & POZZAN, T. 1995. Overexpression of calreticulin increases the Ca2+ capacity of rapidly exchanging Ca2+ stores and reveals aspects of their lumenal microenvironment and function. J Cell Biol, 130, 847-55.

DI BUDUO, C. A., ABBONANTE, V., MARTY, C., MOCCIA, F., RUMI, E., PIETRA, D., SOPRANO, P. M., LIM, D., CATTANEO, D., IURLO, A., GIANELLI, U., BAROSI, G., ROSTI, V., PLO, I., CAZZOLA, M. & BALDUINI, A. 2020. Defective interaction of mutant calreticulin and SOCE in megakaryocytes from patients with myeloproliferative neoplasms. Blood, 135, 133-144.

IBARRA, J., ELBANNA, Y. A., KURYLOWICZ, K., CIBODDO, M., GREENBAUM, H. S., ARELLANO, N. S., RODRIGUEZ, D., EVERS, M., BOCK-HUGHES, A., LIU, C., SMITH, Q., LUTZE, J., BAUMEISTER, J., KALMER, M., OLSCHOK, K., NICHOLSON, B., SILVA, D., MAXWELL, L., DOWGIELEWICZ, J., RUMI, E., PIETRA, D., CASETTI, I. C., CATRICALA, S., KOSCHMIEDER, S., GURBUXANI, S., SCHNEIDER, R. K., OAKES, S. A. & ELF, S. E. 2022. Type I but Not Type II Calreticulin Mutations Activate the IRE1alpha/XBP1 Pathway of the Unfolded Protein Response to Drive Myeloproliferative Neoplasms. Blood Cancer Discov, 3, 298-315.

KAUR, A., VENKATESAN, A., KANDARPA, M., TALPAZ, M. & RAGHAVAN, M. 2024. Lysosomal degradation targets mutant calreticulin and the thrombopoietin receptor in myeloproliferative neoplasms. Blood Adv, 8, 3372-3387.

MERY, L., MESAELI, N., MICHALAK, M., OPAS, M., LEW, D. P. & KRAUSE, K. H. 1996. Overexpression of calreticulin increases intracellular Ca2+ storage and decreases store-operated Ca2+ influx. J Biol Chem, 271, 9332-9.

NAKAMURA, K., ZUPPINI, A., ARNAUDEAU, S., LYNCH, J., AHSAN, I., KRAUSE, R., PAPP, S., DE SMEDT, H., PARYS, J. B., MULLER-ESTERL, W., LEW, D. P., KRAUSE, K. H., DEMAUREX, N., OPAS, M. & MICHALAK, M. 2001. Functional specialization of calreticulin domains. J Cell Biol, 154, 961-72.

PECQUET, C., PAPADOPOULOS, N., BALLIGAND, T., CHACHOUA, I., TISSERAND, A., VERTENOEIL, G., NEDELEC, A., VERTOMMEN, D., ROY, A., MARTY, C., NIVARTHI, H., DEFOUR, J. P., EL-KHOURY, M., HUG, E., MAJOROS, A., XU, E., ZAGRIJTSCHUK, O., FERTIG, T. E., MARTA, D. S., GISSLINGER, H., GISSLINGER, B., SCHALLING, M., CASETTI, I., RUMI, E., PIETRA, D., CAVALLONI, C., ARCAINI, L., CAZZOLA, M., KOMATSU, N., KIHARA, Y., SUNAMI, Y., EDAHIRO, Y., ARAKI, M., LESYK, R., BUXHOFER-AUSCH, V., HEIBL, S., PASQUIER, F., HAVELANGE, V., PLO, I., VAINCHENKER, W., KRALOVICS, R. & CONSTANTINESCU, S. N. 2023. Secreted mutant calreticulins as rogue cytokines in myeloproliferative neoplasms. Blood, 141, 917-929.

PIETRA, D., RUMI, E., FERRETTI, V. V., DI BUDUO, C. A., MILANESI, C., CAVALLONI, C., SANT'ANTONIO, E., ABBONANTE, V., MOCCIA, F., CASETTI, I. C., BELLINI, M., RENNA, M. C., RONCORONI, E., FUGAZZA, E., ASTORI, C., BOVERI, E., ROSTI, V., BAROSI, G., BALDUINI, A. & CAZZOLA, M. 2016. Differential clinical effects of different mutation subtypes in CALR-mutant myeloproliferative neoplasms. Leukemia, 30, 431-8.

SONNICHSEN, B., FULLEKRUG, J., NGUYEN VAN, P., DIEKMANN, W., ROBINSON, D. G. & MIESKES, G. 1994. Retention and retrieval: both mechanisms cooperate to maintain calreticulin in the endoplasmic reticulum. J Cell Sci, 107 (Pt 10), 2705-17.

SUZUKI, J., KANEMARU, K., ISHII, K., OHKURA, M., OKUBO, Y. & IINO, M. 2014. Imaging intraorganellar Ca2+ at subcellular resolution using CEPIA. Nat Commun, 5, 4153.

VENKATESAN, A., GENG, J., KANDARPA, M., WIJEYESAKERE, S. J., BHIDE, A., TALPAZ, M., POGOZHEVA, I. D. & RAGHAVAN, M. 2021. Mechanism of mutant calreticulin-mediated activation of the thrombopoietin receptor in cancers. J Cell Biol, 220, e202009179.

WIJEYESAKERE, S. J., GAFNI, A. A. & RAGHAVAN, M. 2011. Calreticulin is a thermostable protein with distinct structural responses to different divalent cation environments. J Biol Chem, 286, 8771-85.

Version 2

Abstract

Calreticulin (CRT) is an endoplasmic reticulum (ER) chaperone with low affinity calcium binding sites in its C-terminal domain. This region is altered by somatic mutations in the CRT gene (CALR), which drive a subset of myeloproliferative neoplasms (MPN). Perturbations in ER calcium storage and signaling are reported for the MPN type I mutant, CRTDel52, and are linked to disease pathogenesis. Using recombinant CRT proteins, we found similar low affinity calcium binding characteristics for wild-type CRT and CRTDel52, as determined by isothermal titration calorimetry (ITC). Residues 340–349, conserved in both wild-type CRT and CRTDel52, contribute to binding. Furthermore, ER and cytosolic calcium levels and store-operated calcium entry (SOCE) were comparable in CRT knockout (CRT-KO) cells reconstituted with wild-type CRT or CRTDel52. Notably, the CRT-KO induces expression of multiple ER proteins known to contain low affinity calcium binding sites or regulate ER calcium levels. Overall, these findings indicate that CRTDel52 retains at least some low affinity calcium binding capacity and that CRT-deficiency induces compensatory cellular changes that maintain ER calcium homeostasis.

Introduction

Calreticulin (CRT) is an endoplasmic reticulum (ER) luminal protein that plays key roles in the assembly and folding of N-linked glycoproteins. CRT has a single high affinity calcium binding site within a globular domain and a highly acidic C-terminal domain that contains multiple low affinity calcium binding sites (Baksh and Michalak, 1991, Wijeyesakere et al., 2011). As an ER chaperone, CRT is thought to play a critical role in maintaining cellular calcium homeostasis (Venkatesan et al., 2021b, Michalak, 2024). The ER has a relatively high free calcium concentration of 0.5-1 μM through the presence of various low affinity calcium-binding proteins that buffer calcium, whereas cytosolic calcium levels are maintained within a narrow range of around 100 nM (Michalak, 2024). The maintenance of calcium homeostasis within cellular compartments is essential for cell function and survival (Bagur and Hajnoczky, 2017). Early studies have shown that the C-domain of CRT contributes to ER calcium storage, although free ER luminal concentrations are unchanged (Nakamura et al., 2001).

Somatic mutations of exon 9 of the CALR gene including a 52-base pair deletion (Del52; type I mutation) and a 5-base pair insertion (Ins5; type 2 mutation), are driver mutations in some myeloproliferative neoplasms (MPN), specifically in essential thrombocythemia (ET) and primary and secondary myelofibrosis (PMF) (Klampfl et al., 2013, Nangalia et al., 2013). The MPN mutations change the CRT C-domain residues from being predominantly acidic to a high prevalence of basic amino acids and result in a loss of the ER-retention KDEL sequence (Klampfl et al., 2013, Nangalia et al., 2013) (Figure 1). Although all MPN CRT mutants change the sequence and charge of the C-domain, a greater number of acidic residues are retained in MPN type 2 mutant (CRTIns5), compared to the type I mutant (CRTDel52) (Figure 1). The loss of more acidic residues from the C-terminal domain of CRTDel52 has been reported to impair its calcium binding and reduce ER calcium levels (Ibarra et al., 2022). Store-operated calcium entry (SOCE) is a physiological process that allows for the entry of extracellular calcium and ER calcium restoration following ER calcium depletion (Putney, 2011, Zhang et al., 2020). Altered calcium release from the ER and SOCE was reported for MPN type I but not type 2 mutants in cultured megakaryocytes from MPN patients (Pietra et al., 2016). In another study, spontaneous calcium spikes and increased SOCE activation were observed in cultured megakaryocytes from MPN type I and type 2 patients when compared to healthy donors (Di Buduo et al., 2020).

Sequence alignments of the C-terminal residues of murine and human calreticulin (mCRT and hCRT), and the MPN-linked CRT mutants Ins5 and Del52.

Multiple sequence alignment using Clustal Omega was employed to compare C-domain sequences of mature mCRT, hCRT, Del52, and Ins5 proteins. Acidic residues (340 – 349) previously implicated in low affinity calcium binding by mCRT that are shared between wild-type and both CRTDel52 and CRTIns5 are highlighted in green. Additional acidic residues shared just between wild-type CRT and CRTIns5 are shown in red. There are additional acidic residues unique to wild-type CRT. Mature protein numbering is used for the CRT sequences.

Although global calcium binding defects were predicted or qualitatively measured for CRTDel52 (Shivarov et al., 2014, Ibarra et al., 2022) and altered cellular calcium signaling is also reported (Pietra et al., 2016, Di Buduo et al., 2020), it remains unclear whether MPN-linked CRTDel52 has any low affinity calcium binding capacity, given that a few conserved acidic residues are retained in its C-domain (Figure 1). In fact, our previous studies showed that a truncated version of murine wild-type CRT (mCRT) containing the non-mutated part of CRTDel52 and two additional C-terminal residues (mCRT1-351) showed similar low affinity calcium binding characteristics as the wild-type mCRT (Wijeyesakere et al., 2011). On the other hand, further truncation of mCRT (mCRT1-339) was needed to disrupt calcium binding (Wijeyesakere et al., 2011, Wijeyesakere et al., 2016). The goal of the current study was to quantitatively measure low affinity calcium binding by CRTDel52 compared with wild-type human CRT and also quantify ER and cytosolic calcium levels and calcium responsiveness in cells expressing wild-type CRT or CRTDel52.

Results

MPN-linked CRTDel52 retained at least some low affinity calcium binding sites

Previous studies (Wijeyesakere et al., 2011) have implicated residues 340 to 351 of murine CRT, which are identical in wild-type human CRT and fully conserved with the 340 to 349 residues of CRTDel52, to be involved in low affinity calcium binding. Further truncation of the C-domain to the residue 339 (mCRT1-339) led to the loss of calcium binding measured by ITC (Wijeyesakere et al., 2011). Figure 1 shows a sequence alignment of murine CRT, human CRT, CRTIns5, and CRTDel52 proteins and indicates that the residues previously implicated in low affinity calcium binding are also shared between wild-type CRT and CRTDel52. Recombinant his and GB1 tagged purified proteins were used to compare calcium binding by human CRT and its variants. Chromatograms for human wild-type CRT, CRTDel52, CRT(1-339) and CRT(1-351) are shown in Figure 2A. SDS-PAGE gels corresponding to the purified proteins are shown in Figure 2B. Intact protein mass spectrometry for CRTDel52 indicated 35-42 amino acid truncations relative to the full protein sequence for different preparations, possibly from C-terminal protein degradation in E. coli. The truncated proteins still include acidic residues (340–351) previously implicated in low affinity calcium binding by murine CRT that are shared between wild-type and CRTDel52 (Figure 1).

Purification of recombinant human wild-type CRT, CRTDel52, and C-domain truncation mutants.

A) Representative chromatograms of recombinant GB1-and his-tagged human CRTWT (hCRTWT), CRTDel52, and the C-domain truncation mutants hCRT (1-351) and hCRT (1-339) purified by gel exclusion chromatography after expression in E. coli and initial purification using a nickel resin. B) Representative Coomassie blue-stained SDS-PAGE gels of fractions corresponding to the monomer peak (those contained within the region indicated by the double headed arrows on the chromatogram) of purified proteins that were used for calcium-binding measurements.

CRTIns5 was not purified in sufficient quantities needed for ITC experiments and was not further investigated in this study.

For the ITC measurements, calcium titrations were performed after pre-blocking the high affinity calcium binding sites with 50-100 µM CaCl2. In a previous study (Wijeyesakere et al., 2011), we showed that at a starting calcium concentration of 0 mM and with CaCl2 injections to a final concentration of 70-80 μM, the measured KD value was 16.6 μM for calcium binding to wild type murine CRT, (which has ∼95% sequence identity with human CRT), corresponding to the high affinity site. On the other hand, at a starting calcium concentration of 50 μM and CaCl2 injections (at 33 μM each) to a final concentration of 700-850 μM, the measured KD value for calcium binding to wild type murine CRT was 590 μM (corresponding to the low affinity sites). We did not measure the high affinity sites under the latter conditions (Wijeyesakere et al., 2011). In the present experiments, at starting CaCl2 concentrations of 50-100 μM and 25 sequential 10µL injections of 76 μM CaCl2 to a final concentration of 1650-1700 μM, similar binding profiles and KD values in the 400-1000 μM range were measured with wild-type human CRT, CRTDel52, and the CRT(1-351) mutant (Figures 3A, C and D and Table 1). However, consistent with previous studies with mCRT(1–339), endothermic calcium binding by human CRT(1–339) was of lower magnitude compared to other constructs, with a KD value that was too low to be reliably fit, suggesting the loss of low affinity calcium binding sites in this mutant (Figure 3B and Table 1). There are exothermic peaks observed with human CRT(1–339) at the beginning of the calcium titration, which may be attributed to altered low affinity sites retained within the mutant or reduced affinity of the high affinity sites. We concluded that CRTDel52 retains at least some low affinity calcium binding sites that are measurable by ITC.

Calcium binding to the low affinity sites of purified human CRT proteins.

Low affinity calcium binding by purified hCRTWT, CRTDel52, and the indicated truncation mutants was measured using isothermal titration calorimetry. A-D) Representative isotherm and fitting curves of calcium binding to the low affinity sites of the wild-type and mutant CRT proteins as indicated, measured using isothermal titration calorimetry (ITC). Prior to titration, the proteins were incubated with 50-100 µM CaCl2 to block the high affinity calcium binding sites. At 37°C, 10 mM CaCl2 was titrated against 100 µM protein. The heat change was measured for 25 injections, and the data were curve-fitted to calculate the dissociation constant, enthalpy, and binding stoichiometry using NanoAnalyze software. Data replicates and statistical analyses are specified in Table 1.

Similar ER calcium levels in CRT-KO HEK cells reconstituted with wild-type CRT or CRTDel52

While the CRTDel52 mutant retains some ability to bind calcium (Figure 3 and Table 1), the loss of the ER-retention KDEL sequence induced by the frameshift mutation, which results in the secretion of the protein from the ER (Araki et al., 2016, Chachoua et al., 2016, Han et al., 2016, Arshad and Cresswell, 2018, Liu et al., 2020, Masubuchi et al., 2020, Venkatesan et al., 2021a, Pecquet et al., 2023) might be expected to cause ER calcium insufficiency in cells expressing CRTDel52 compared to wild-type CRT. We investigated ER calcium signals in CRT knock-out HEK293T cells generated by CRISPR/Cas9 gene editing. A clonal KO line was generated and subsequently reconstituted with constructs encoding wild-type CRT, CRTDel52, or CRTDel52 with a KDEL sequence added at the C-terminus (CRTDel52-KDEL). Protein expression was assessed by immunoblotting with anti-CRT(N), which detects an N-domain epitope present in all CRT constructs, and anti-CRT(C-mut), which was specifically raised against the mutant CRT C-terminus (Venkatesan et al., 2021a). While the recognition with anti-CRT (N) was very weak in lysates of cells expressing CRTDel52, which is secreted, CRTDel52-KDEL is readily detectable with both antibodies, consistent with prior studies demonstrating the contributions of the KDEL sequence to the ER retention of CRT and its mutants, including in HEK cells (Arshad and Cresswell, 2018, Sonnichsen et al., 1994). Both wild-type CRT and CRTDel52-KDEL were reconstituted to higher levels than in the parental cells (Figure 4A, lanes 1 and 2 compared to 7-8 and 11-12).

Flow cytometry-based ER calcium measurements indicate similar ER calcium levels in CRT-KO HEK293T cells and those reconstituted with wild-type CRT and CRTDel52.

A) CRISPR-Cas9 editing of HEK cells using lentiviral constructs with calreticulin-targeting gRNA (CRT-KO) or empty vector (HEK-Vec). CRT-KO HEK cells were subsequently reconstituted with wild-type human CRT, CRTDel52, or CRTDel52 with a C-terminal KDEL sequence as indicated. Representative blots showing expression levels of CRT in the indicated HEK cells. Cell lysates were subjected to SDS-PAGE followed by immunoblotting. Wild-type CRT, mutant CRT, and GAPDH were detected with anti-CRT (N), anti-CRT C-(mut), and anti-GAPDH antibodies, respectively. B) Gating strategy for flow cytometry. HEK-Vec cells transiently transfected with the vector encoding the ratiometric GEM-CEPIA1er probe for 72 hours were used to assess ER calcium levels by flow cytometry based on the measurements of fluorescence emission signals for the Ca2+-bound or unbound probe in the Pacific Blue and Amcyan channels, respectively following excitation with the 405 nm (violet) laser. Transfected cells were gated based on forward (FSC-A) and side scatter (SSC-A), then on 7AAD-negative (live) cells, and further on the cells positive for the calcium-bound and unbound probe signals. Representative dot plots of changes in fluorescence intensities in response to treatment with thapsigargin (Tg), a SERCA inhibitor, before (0 mins) or after (10 mins) treatment, are shown. C) Representative histograms showing changes to Ca2+-bound and unbound signals at different time points following Tg treatment of cells. Mean fluorescence intensities (MFI) for Ca2+-bound and unbound signals were used to calculate the bound/unbound ratios at different time points. D and E) The basal free ER calcium signals were measured as bound/unbound MFIs ratios from the corresponding histograms of CRT-KO or HEK-Vec cells (D, left panel) or CRT-KO cells reconstituted with the indicated CRT constructs (D, right panel). The fractional changes in ER calcium signals in response to Tg were measured in the indicated cells (E). The fractional change ratios were calculated as (R0-R10/R0)x100. Statistical significance was determined using GraphPad Prism and based on paired t-tests (left panels of D and E) or repeated measures one-way ANOVA analyses with Tukey’s test (right panels of D and E). ns, not significant; *** P value < 0.001. All measurements were undertaken in the presence of 2 mM extracellular Ca2+. Data shown in D and E are based on 7-15 independent experiments as indicated by the individual data points within the graphs. GAPDH, glyceraldehyde-3-phosphate dehydrogenase; SDS-PAGE, sodium dodecyl sulfate-polyacrylamide gel electrophoresis.

The KO and reconstituted cells were further transfected with pCIS-GEM-CEPIA1er, which encodes a ratiometric ER calcium probe (Suzuki et al., 2014). Emission signals for calcium-bound (peak ∼ 460 nm) and unbound (peak ∼ 510 nm) probe were measured using flow cytometry in the Pacific Blue and AmCyan channels, respectively, following excitation with the 405 nm (violet) laser. Flow cytometric analysis was undertaken in a buffer containing 2 mM extracellular Ca2+ (Figures 4B and 4C). There was a significant decrease in basal free ER calcium, measured by the bound/unbound ratios in CRT-KO cells compared to the parental cells, (Figure 4D, left panel). ER calcium levels were partially restored, although non-significantly, when the CRT-KO cells were reconstituted with wild-type CRT. CRT-KO cells reconstituted with CRTDel52 and CRTDel52-KDEL showed small but non-significant decreases in ER calcium levels compared to those reconstituted with wild-type CRT (Figure 4D, right panel). Changes to ER calcium levels in response to thapsigargin (Tg), a SERCA inhibitor, were similar in the parental vs. CRT-KO cells, and in the CRT-KO cells compared with those reconstituted with different CRT proteins (Figure 4E).

Spectrofluorimetry was used to verify the maintenance of ER calcium signals in the different HEK cell lines. Cells were again transfected with the pCIS-GEM-CEPIA1er probe and spectra were recorded in a buffer containing 2 mM Ca2+, following excitation at 390 nm. Representative spectra are shown in Figure 5A. Consistent with the flow cytometric measurements, there were no significant differences in the basal ER calcium levels between CRT-KO cells and those reconstituted with wild-type CRT or CRTDel52 (Figure 5B). There were also similar changes to ER calcium levels in response to Tg treatment in the CRT-KO cells compared to cells reconstituted with either wild-type CRT or CRTDel52 (Figure 5C).

Fluorimetry-based ER calcium measurements indicate similar ER calcium levels in CRT-KO HEK293T cells and those reconstituted with wild-type CRT and CRTDel52.

HEK cells transiently transfected with a vector encoding the ratiometric GEM-CEPIA1er probe for 72 hours were used to assess ER calcium levels in the presence of 2 mM extracellular Ca2+. Fluorescence emission spectra were recorded following excitation at 390 nm. A) Representative emission spectra show changes to Ca2+-bound (∼460 nm) and unbound (∼510 nm) peaks at different time points following Tg treatment of cells. The spectrum in blue was obtained with untransfected HEK cells. B) Basal Ca2+-bound (∼460 nm)/unbound (∼510 nm) fluorescence intensity (FI) ratios were calculated for the indicated HEK cell lines before Tg addition. C) The fractional changes to the Bound (∼460 nm)/Unbound (∼510 nm) florescence intensity ratios at 15 mins relative to baseline were calculated as (R0-R15/R0)x100. Statistical significance was determined by repeated measures one-way ANOVA analyses with Tukey’s test using Graph Prism. Data shown are based on 4 independent experiments, indicated by the individual data points within the graphs. ns, not significant

Similar cytosolic calcium and SOCE signals in a CRT-KO megakaryocyte cell line reconstituted with wild-type CRT or CRTDel52

MPN CRT mutations cause thrombopoietin-independent differentiation of megakaryocytes to platelets via the thrombopoietin receptor (MPL) (Araki et al., 2016, Chachoua et al., 2016, Marty et al., 2016, Li et al., 2018, Masubuchi et al., 2020). We, therefore, used a megakaryoblast cell line, MEG-01, to assess the role of CRT deficiency and the CRTDel52 mutant on basal cytosolic calcium levels and SERCA inhibition-induced calcium influx into the cytosol. As with the HEK cells (Figure 4), we created a stable CRT knockout clone of MEG-01 cells by CRISPR/Cas9 editing for cytosolic calcium measurements. The cells were subsequently reconstituted with wild-type CRT or CRTDel52. Representative immunoblots are shown to demonstrate the successful knockout and reconstitution (Figure 6A). Wild-type CRT reconstitution in MEG-01 CRT-KO cells resulted in expression levels higher than those in the parental cells (Figure 6A, lanes 7 and 8 compared with lanes 1 and 2). In the presence of 2 mM Ca2+ in the media, we observed no significant differences in the basal cytosolic calcium levels or CPA-induced calcium release from ER in CRT-KO cells compared to the parental MEG-01 cells (Figure 6C and 6D). Additionally, no significant differences in basal cytosolic calcium were observed in CRT-KO cells compared to those reconstituted with wild-type CRT or CRTDel52 (Figure 6E). Furthermore, cytosolic calcium levels in CRT-KO and the two reconstituted cell lines were not statistically different following CPA treatments.

Calcium imaging studies indicate similar cytosolic calcium levels in the megakaryoblastic (MEG-01)cells with CRT-KO and those reconstituted with wild-type CRT or CRTDel52

A) CRISPR-Cas9 editing of MEG-01 cells using lentiviral constructs encoding calreticulin-targeting gRNA (CRT-KO) or empty vector (Vec). CRT-KO MEG-01 cells were subsequently reconstituted with viral constructs encoding wild-type human CRT or CRTDel52 or transduced with empty vector (Vec) as indicated in the Figure panels. Representative immunoblots of CRT expression in the indicated cells detected by the anti-CRT(N) and anti-CRT(C-mut) antibodies. B-F) Cells in 2 mM extracellular Ca2+ were loaded with the Fura-2/AM dye. Following excitation at 340 or 380 nm and recordings of emission at 510 nm, the emission ratios (340/380) are calculated. B) Representative calcium imaging data from indicated cells in the presence of 2 mM extracellular Ca2+ and following ER calcium depletion with SERCA inhibitor, cyclopiazonic acid (CPA). C and E) Basal cytosolic calcium levels are plotted as the mean 340/380 ratios. Data were normalized relative to the Vec signals. D and F) Mean fractional changes to cytosolic calcium levels calculated as (Rpeak-Rbaseline/Rbaseline)x100 and total calcium release determined as the area under the curve (AUC), are plotted. Data shown in C-F are based on 4 or 5 independent experiments with measurements from averaged 50-100 cells per experiment. Statistical analysis was performed using paired t-tests (C and D) or repeated measures one-way ANOVA analyses with Tukey’s test (E and F). Statistical significance was determined based on p-value ≤ 0.05 using GraphPad Prism. ns, not significant.

CALR mutations are generally found in BCR-ABL1-negative MPN (Klampfl et al., 2013, Nangalia et al., 2013), whereas MEG-01 cells have the BCR-ABL1 translocation (Ogura et al., 1985), which induces constitutive tyrosine kinase activity. This could influence intracellular calcium homeostasis (Cabanas et al., 2018). To eliminate possible effects of BCR-ABL1 expression on the measured responses (Figure 6), cytosolic calcium signals in the various MEG-01 cell lines were compared with and without treatment of cells with the tyrosine kinase inhibitor, Imatinib. We initially verified the effect of Imatinib on MEG-01 cells by assessing the expression of STAT5 and the formation of phospho-STAT5 using immunoblots. These analyses showed significantly reduced or abrogated phospho-STAT5 expression upon Imatinib treatment (Figure 6-figure supplement 1). Similar to the findings in the absence of Imatinib (Figures 6 and Figure 6-figure supplement 2A, left panel), there were no significant differences in the basal cytosolic calcium levels in Imatinib-treated CRT-KO cells compared to those reconstituted with wild-type CRT and CRTDel52 (Figure 6-figure supplement 2B, left panel). Likewise, CPA-induced calcium release from ER was generally similar in CRT-KO cells compared to cells reconstituted with wild-type CRT or CRTDel52 both in imatinib-treated and untreated cells (Figure 6-figure supplements 2A and 2B, middle and right panels).

SOCE is an important process for restoring ER calcium and maintaining intracellular calcium levels. In contrast to cytosolic calcium signals in the presence of 2 mM Ca2+ in the media, in calcium-free saline, CRT-KO cells exhibited a small but significant increase in basal cytosolic calcium levels relative to the parental cells (Figure 7A, left panel and Figure 7B). Upon ER calcium depletion with CPA, the mean fractional increase in the cytosolic calcium measured in the Ca2+-free saline did not differ significantly in parental cells vs. CRT-KO cells. The areas under the curve were also not different for the two cell lines (Figures 7A and 7C). However, consistent with the higher basal cytosolic calcium measured in CRT-KO cells in calcium-free saline, the measured SOCE signal (following the addition of 2 mM Ca2+ and CPA) was significantly reduced in CRT-KO cells compared to parental cells (Figure 7D). Reconstitution of the CRT-KO cells with either wild-type CRT or CRTDel52 tended to reduce basal cytosolic calcium and induce SOCE (Figure 7A, right panel, 7E and 7G), but both trends were non-significant compared to CRT-KO cells. Importantly, there were no significant differences between the wild-type CRT and CRTDel52 reconstituted cells, in either their basal or CPA-induced cytosolic calcium signals or in SOCE (Figures 7E-G). In calcium-free saline, cytosolic calcium signals and SOCE induction were generally similar in CRT-KO cells as in the wild-type CRT and CRTDel52 reconstituted cells, both in the presence and absence of imatinib (Figure 7-figure supplement 1).

Calcium imaging studies indicate similar cytosolic calcium levels and SOCE in MEG-01 cells with CRT-KO and those reconstituted with wild-type CRT or CRTDel52.

Cells were loaded with Fura-2/AM dye. Following excitation at 340 or 380 nm and recordings of emission at 510 nm, the emission ratios (340/380) were calculated. A) Representative calcium imaging data from cells in the absence of extracellular calcium, followed by the addition of CPA, and subsequent addition of extracellular calcium and CPA as indicated. B and E) Basal cytosolic calcium levels are plotted as the mean 340/380 ratios prior to CPA addition. Data are normalized relative to the signals of cells transduced with empty vector (Vec). C and F) Fractional changes to cytosolic calcium calculated as (Rpeak-Rbaseline/ Rbaseline)x100 (left panel) and total calcium release measured as the area under the curve (AUC) (right panel) are plotted for the indicated cells, following ER calcium depletion with CPA. D and G) Fractional changes to cytosolic calcium following CPA and extracellular calcium additions (2 mM) are calculated as (Rpeak-Rbaseline/ Rbaseline)x100, where Rbaseline and Rpeak were measured over 1-minute intervals. Data shown are based on 8-10 independent experiments, and an average of over 20-100 cells in each experiment. Comparisons were performed by paired t-test (panels B-D) and one-way ANOVA analysis (panels E-G) using GraphPad Prism. Statistical significance is based on a p-value ≤ 0.05. ns, not significant.

CRT deficiency induced numerous transcriptional changes to genes within the calcium signaling pathway

As noted in Figures 4 and 7, some of the measured differences between CRT-KO and parental cells are not fully reversed by reconstitution with wild-type CRT, although the trends are maintained. We predicted that there may be other indirect effects of CRT-KO and reconstitution that affect cellular calcium levels and calcium signaling that could explain some of these measured differences. As proof of concept, we undertook bulk RNA-sequencing to determine differential gene expression in MEG-01 parental vs. CRT-KO cells. There were 3078 upregulated and 1790 downregulated genes. The volcano plot and heat map show identified genes that were differentially expressed in CRT-KO MEG-01 compared to parental cells (Figure 8A). Based on pathway analysis, numerous genes involved in cellular calcium signaling were identified. Of specific note is the upregulation of GRP94 encoded by the HSP90B1 gene, BiP encoded by HSPA5 gene, and PDIA6 encoded by PDIA6 gene, which are known to have low affinity calcium binding sites (Lievremont et al., 1997, Macer and Koch, 1988, Biswas et al., 2007, Marzec et al., 2012, Okumura et al., 2021) (Figures 8A and 8B). Other studies have reported similar upregulation of these genes in CRT-KO (Tang et al., 2025) as well as in heterozygous CRTDel52 knock-in cells (Figure 8C) (Fosselteder et al., 2023). Upregulation of these proteins may serve a compensatory role in buffering calcium in CRT-KO cells. Additionally, PDIA3, which is shown to regulate the activities of stromal interaction molecule 1 (STIM1) (Prins et al., 2011) and Sarco/endoplasmic reticulum Ca2+ ATPase (SERCA2b) (Li and Camacho, 2004), was consistently upregulated in all the studies.

Transcriptional changes to the expression of genes encoding ER calcium binding/regulating proteins in CRT-KO MEG-01 cells.

A) RNA sequencing analysis was used to identify differentially expressed genes in Meg-01 CRT-KO cells in comparison to parental MEG-01 cells. Volcano plot shows significantly upregulated (blue) and downregulated genes (red) genes identified by RNA sequencing. Genes related to ER calcium binding and homeostasis have been highlighted and labeled. The differentially expressed genes were selected as those with a log2 fold-change of <-0.49 or >0.49 and p values <0.05. B) Heat map shows the differential expression of genes encoding ER resident calcium binding proteins or ER calcium regulators that were identified by RNA sequencing in CRT-KO Meg-01 cells compared to parental cells (n=3 biological replicates). C) Comparison of expression ratio of HSPA5, HSP90B1, PDIA3 and PDIA6 in CRT-KO MEG-01 cells (this study) or murine lung cancer cells (Tang et al., 2025), heterozygous CRTDel52 knock-in (Del52-KI) cells (Fosselteder et al., 2023) normalized to their expression in respective control cell lines expressing wild type CRT. The normalized expression of various genes identified in each dataset was calculated as a ratio of expression in CRT-KO, CRTDel52 knock-in or CRTDel52 cells relative to the expression in wild-type CRT-expressing cells. Multiple unpaired t-tests were used to determine statistical significance with the Benjamini-Hochberg FDR correction (p-adjusted) <0.05. D-G) Left and middle panels: Relative expression of mRNA transcripts of HSPA5 (n=8-11), HSP90B1 (n=7), PDIA6 (n=6) and PDIA3 (n=6) genes in the indicated MEG-01 cells determined by RT-qPCR using specific gene primers. Relative gene expression was calculated by normalizing the target gene’s cycle threshold (Ct) to a single or multiple endogenous controls (ACTB, GAPDH and HPRT1) which was used for comparison across different cell lines as indicated. Data show 6 - 11 independent runs with duplicate measurements for each run. Student t-test was used to determine the statistical significance between two groups while multiple group comparison was performed by one-way ANOVA analysis using GraphPad Prism. Statistical significance is based on a p-value ≤ 0.05. ns, not significant. Right panels: Representative immunoblots of BiP, GRP94, PDIA6 and PDIA3 expression in the indicated cells detected by specific antibodies.

We further used real-time PCR with target gene primers (Supplementary Table 2) to validate the expression of the key ER resident proteins identified from the RNA-sequencing data which have roles in calcium binding and regulation of ER calcium homeostasis. Consistent with the RNA-seq data, the expression of HSPA5, HSP90B1, PDIA6 and PDIA3 mRNA transcripts was significantly increased in CRT-KO MEG-01 compared to parental cells. Reconstitution of CRT-KO cells with wild-type CRT but not CRTDel52 showed a significant decrease in the expression of HSPA5, HSP90B1, PDIA6 and PDIA3 (Figures 8D-G). Protein levels of BiP, GRP94, PDIA6 and PDIA3 were also assessed in the indicated cells by immunoblots using specific antibodies. Representative immunoblots show increased expression of BiP (Figure 8D, right panel), GRP94 (Figure 8E, right panel), PDIA6 (Figure 8F, right panel) and PDIA3 (Figure 8G, right panel) in CRT-KO cells compared to parental MEG-01 cells as well as increased expression in CRT-KO-Vec cells compared to CRT-KO cells reconstituted with wild type CRT. The effects of CRTDel52 in reducing expression were generally less robust than wild type CRT.

Discussion

Calreticulin is a ubiquitously expressed ER luminal calcium-binding chaperone, with a role in ER calcium storage (Venkatesan et al., 2021b, Michalak, 2024). A number of prior studies have examined the effects of CRT deficiency and over-expression upon cytosolic and ER calcium signals and SOCE (Arnaudeau et al., 2002, Bastianutto et al., 1995, Mery et al., 1996, Nakamura et al., 2001). CRT deficiency in mouse fibroblasts did not affect free ER calcium levels relative to wild type cells (Nakamura et al., 2001). Previous studies using dye-based assays reported impaired calcium binding by the MPN-associated CRTDel52 mutant (Ibarra et al., 2022). Yet direct and more quantitative analyses of its calcium-binding properties have been lacking. In this study, we investigated low affinity calcium binding by wild-type human CRT, CRTDel52, and the truncation mutants CRT(1-351) and CRT(1-339) using ITC (Figure 3 and Table 1). Our studies show that wild-type CRT, CRTDel52, and CRT(1-351) retain low affinity calcium binding sites, exhibiting average KD values of 574.24, 938.82, and 444.34 µM, respectively. On the other hand, low affinity calcium binding by CRT(1-339) could not be reliably fit. This is the first quantitative characterization of the low affinity calcium binding profiles of wild-type human CRT, CRTDel52, and functionally informative truncation mutants.

Although CRTDel52 retains at least partial low affinity calcium binding capability, loss of its KDEL retention sequence (Figure 1) could still disrupt ER calcium storage and signaling. Interestingly, the cellular assays reveal that ER and cytosolic calcium levels, as well as SOCE, are similar between CRT-KO cells and those reconstituted with either wild-type CRT or CRTDel52 (Figures 4-7). One study has reported decreased ER calcium levels in a human osteosarcoma cell line transiently transfected to over-express a CRT type I mutant, compared to those over-expressing wild-type CRT or a type 2 mutant, which was linked to unfolded protein response (UPR) induction in type I mutant-expressing cells (Ibarra et al., 2022). The same study suggested the contributions of the P and C-domains of wild-type CRT in restoring ER calcium levels in cells expressing CRTDel52 (Ibarra et al., 2022). The conditions of those experiments differ from the studies described in our study which uses a CRT-KO background with wild type CRT or CRTDel52 reconstitutions to simulate homozygous MPN type I CALR mutations. Achieving endogenous expression levels of wild type CRT is a limitation of our study as in many other studies examining the effects of CRT on cellular calcium signaling. Despite this, our findings are consistent with the theoretical prediction that any excess free ER calcium generated in cells lacking CRT or sufficient ER calcium binding sites would rapidly diffuse out of the ER and be pumped out of the cell.

Related to cytosolic calcium levels and SOCE activation, consistent with our findings (Figures 7G and Figure 7-figure supplement 1), a previous study found no differences in cytosolic calcium levels or SOCE induction in myeloid progenitor 32D cells expressing wild-type CRT and CRTDel52 both under conditions of cytokine deprivation and thrombopoietin (TPO) stimulation (Bhuria et al., 2024). On the other hand, studies with cultured megakaryocytes from MPN patients with CALR mutations have reported cytosolic calcium abnormalities compared to healthy control cells (Di Buduo et al., 2020, Pietra et al., 2016). In one study, cultured megakaryocytes from patients with type I but not type II CALR mutations were found to have high cytosolic calcium and SOCE activation in cultured megakaryocytes (Pietra et al., 2016). In another study, a larger number of cultured megakaryocytes from MPN patients with type I or type II CALR mutations displayed spontaneous cytosolic calcium spikes and reduced CRT interactions with stromal interaction molecule I (STIM-1) (reflective of SOCE) compared with megakaryocytes from healthy donors or patients with the JAK2V617F mutation (Di Buduo et al., 2020). Thus, there are differences in the findings of the two studies with primary cultured megakaryocytes, the first of which reported increased SOCE activation in cultured megakaryocytes from patients with type I but not type 2 CALR mutations, and the second of which reported increased frequency of cytosolic calcium spikes and reduced STIM-1-CRT binding in cultured megakaryocytes from patients with both type I and type 2 CALR mutations (Pietra et al., 2016, Di Buduo et al., 2020). In the absence of engineered controls, sample to sample heterogeneities in primary cells make it difficult to attribute any measured differences to direct rather than indirect effect of CRT mutations on cellular calcium signaling.

We found that CRT-KO HEK cells displayed lower ER calcium compared with the parental HEK cells (Figure 4D, left panel) and also that CRT-KO MEG-01 cells displayed increased cytosolic calcium in the absence of extracellular calcium (Figure 7B). However, reconstitutions with wild-type CRT did not reverse these effects (Figure 4D, right panel and Figure 7E). Transcriptomic analysis identified significant alterations in the expression of genes related to calcium storage and signaling between parental and CRT-KO MEG-01 cells (Figure 8). These changes highlight adaptive or compensatory pathways induced by CRT-KO and emphasize that, when comparing wild-type and CRT-KO cell lines for subcellular calcium levels or responses to SERCA inhibitors, the measured effects might not be based solely on the abilities of CRT to maintain ER and cytosolic Ca2+ levels via its calcium binding, storage and buffering capacity. Rather, many other compensatory cellular changes could affect subcellular calcium levels and signaling. In this study, various ER proteins with known calcium binding activities were found to be upregulated in CRT-KO cells (Figure 8A and 8B). GRP94 (Macer and Koch, 1988, Biswas et al., 2007, Marzec et al., 2012), BiP (Lievremont et al., 1997) and our unpublished findings), and PDIA6 (Okumura et al., 2021) are each known to contain low affinity calcium binding sites. CRT-KO cells have significantly induced expression of the genes encoding these proteins. Wild type CRT reconstitution into the CRT-KO cells significantly reduced the expression of these ER proteins, whereas CRTDel52 reconstitution only partly and non-significantly reduced the expression of HSP90B1 and PDIA6 and did not reduce the expression of HSPA5 (Figures 8D-F). GRP94, BiP and PDIA6 have functions in ER protein folding and quality control (Hendershot et al., 2024, Okumura et al., 2021). Related to the mechanisms of induction of calcium binding proteins, it is known that CRTDel52 has impaired chaperone activity towards its substrates compared with wild type CRT (Arshad and Cresswell, 2018, Schurch et al., 2022, Kaur et al., 2025). This could account for some of the measured differences between wild type CRT and CRTDel52 in altering the expression of the ER calcium binding proteins that also function as chaperones and folding factors. However, the induction of GRP94, BiP and PDIA6 in knock-in cells with heterozygous expression of wild type CRT and CRTDel52 (Fosselteder et al., 2023), where the wild type CRT appears sufficient for protein folding activities (Kaur et al., 2025, Schurch et al., 2022), points to compensatory effects relevant to calcium signaling. Regardless of the precise mechanisms relevant to induction of the calcium binding proteins, their combined expression, despite protein compositional differences between the KO and reconstituted cells, appears sufficient to maintain ER calcium levels (Figure 4D, right panel), calcium release from the ER (Figure 4E, right panel, and 6F) and cytosolic calcium levels (Figure 6E).

PDIA3 (ERp57) forms a complex with CRT and interacts with and regulates the activities of the ubiquitously expressed SERCA2b (Li and Camacho, 2004). Its expression is also induced by the CRT-KO and suppressed by wild type CRT expression but non-significantly by CRTDel52 expression. Altered SERCA2b activity could also affect basal ER calcium levels, contributing to similarities between CRT-KO, and wild type CRT or CRTDel52 reconstituted cells.

Collectively, our findings reveal that CRTDel52 retains some ability to bind calcium at low affinity sites. CRT-KO cells reconstituted with wild-type or CRTDel52 have comparable ER and cytosolic calcium levels and store-operated calcium entry (SOCE). CRT-KO induces broad changes in ER calcium binding protein expression as well as in the expression of ERp57. Overall, we can conclude that a number of factors that contribute to the maintenance of ER calcium levels and regulation of calcium transport into the ER are altered in CRT-KO cells. Thus, any disease-related changes in calcium signaling and homeostasis are likely multifactorial, encompassing transcriptional regulation and altered expression of critical mediators of calcium storage, transport and signaling pathways. These studies provide a more nuanced understanding of CRT’s role in cellular calcium signaling and may inform future therapeutic strategies targeting calcium signaling pathways in hematological malignancies.

Material and Methods

DNA constructs

Bacterial expression vectors

Wild-type human CRT (accession number P27797) and CRTDel52 described earlier (Venkatesan et al., 2021a) was expressed as a fusion protein with the B1 domain of streptococcal protein G (GB1) (Huth et al., 1997) and hexa-histidine tag using the pGB1 vector via ligation-independent cloning, as described earlier (Del Cid et al., 2010, DelProposto et al., 2009). Using the pGB1-hCALR plasmid as a template, constructs for expressing his-tagged CRT (1-351) and CRT (1-339) truncation mutants were generated using QuikChange Site-Directed Mutagenesis Kit (Agilent Technologies) following the manufacturer’s protocol and using primers indicated in the Supplemental Table 1.

CRT knockdown

The sgRNA for CRT knockdown was designed to target the GGCCACAGATGTCGGGACCT sequence found at the boundary of intron 3 and exon 4 of the CALR gene (Mohan et al., 2020). The primers for cloning of sgRNA into pLentiCRISPRv2-BLAST (pLB) vector are given in the Supplemental Table 1.

pSIP vectors for CRT expression in human cells

The pMSCV constructs used for expression of wild-type CRT and the CRTDel52 described earlier (Venkatesan et al., 2021a) were used as templates to subclone wild-type CALR and CALRDel52 cDNA sequences into the pSIP-ZsGreen (pLVX-SFFV-IRES-Puro) vector (Bashirova et al., 2020) at the EcoRI and NotI sites using primers given in the Supplemental Table1. The cloning of CALR genes at the EcoRI and NotI sites within the pSIP-ZsGreen vector resulted in the removal of a fragment carrying ZsGreen so that the transduced cells did not express ZsGreen.

gRNA resistant versions for reconstitution of CALR expression in KO cells

Silent mutations were introduced into the target sequence for sgRNA in the pSIP-CALRWT and pSIP-CALRDel52 constructs by site-directed mutagenesis to make them resistant to Cas9-mediated editing. The QuikChange II XL Site-Directed Mutagenesis Kit (Agilent) was used along with primers listed in the Supplementary Table 1. The PCR reaction was set up according to the manufacturer’s protocol, followed by DpnI treatment for 1 h at 37°C to digest the parental DNA strand. The DpnI-treated product was used to transform XL10-Gold ultracompetent cells. Two sets of primers were used to introduce a total of six silent mutations: C1QF and C1QR primers were used in the first round, and C2QF and C2QR primers were used to introduce further mutations into the DNA templates generated from the first round of mutagenesis.

Addition of a KDEL sequence into CRTDel52

The pSIP-CALRDel52-KDEL construct containing a C-terminal KDEL motif was generated using sgRNA-resistant pSIP-CALRDel52 construct as a template and primers given in Supplemental Table 1. The amplicons were cloned at the EcoRI and NotI sites in the pSIP vector. The cloning of CALR genes at the EcoRI and NotI sites within the pSIP-ZsGreen vector resulted in the removal of a fragment carrying ZsGreen so that the transduced cells did not express ZsGreen.

Cell culture, viral transduction and reconstitution of the expression of CRT mutants

Megakaryoblastic leukemia cell line (MEG-01) from ATCC and Human embryonic kidney (HEK) 293T cell lines were cultured and maintained in RPMI 1640 and DMEM media, respectively, supplemented with 10% heat inactivated fetal bovine serum (FBS, Gibco), antibiotic/antimycotic solution containing penicillin, streptomycin and amphotericin (15240062) respectively at 37°C, 5% CO2 and 95% humidity. Wild-type CRT was knocked out in MEG-01 cells and HEK293T cells (CRT-KO) using Cas9 and single guide RNA (sgRNA) targeting GGCCACAGATGTCGGGACCT sequence within the CALR gene. For CRISPR/Cas9 mediated CRT knockdown, the HEK293T cells were first transfected with the lentiviral construct containing CRT-targeting sgRNA (pLB-CALR sgRNA) or the empty pLB vector along with the psPAX2 and MD2.G vectors using Lipfectamine™ LTX with PLUS™ reagent (Invitrogen) to make lentivirus. The lentivirus-containing media were harvested, filtered with 0.45 µm syringe filters and added onto the target cells (MEG-01 or HEK293T cells) for infection. The MEG-01 cells were infected by spinoculation at 2500 rpm for 2 hours at room temperature in the presence of polybrene (8 µg/mL), followed by replacement of 80% of virus-containing media with fresh complete media.

For HEK293T cells, the harvested viral supernatant was added to HEK 293T cells and incubated for 2 hours at 37°C, after which 80% of the viral suspension was replaced with complete growth media. Virus-containing media was completely replaced with fresh growth media 24 hours after infection, and the drug selection started with 2.5 µg/mL blasticidin 48 hours post-infection. Following drug selection, the CRT-KO cells were plated at a dilution of one cell per well in a 96-well plate to isolate single-cell clones.

The selected single-cell clones of CRT-KO MEG-01 and HEK293T cells were subsequently reconstituted with wild-type human CRT, CRTDel52, or CRTDel52 with a C-terminal KDEL sequence using the respective sgRNA-resistant pSIP constructs and the same method of transduction explained above. The cells transduced with the empty pSIP vector without ZsGreen (Supplemental Table 1) were also generated to serve as control cells. Cells expressing CRT proteins or those containing empty pSIP vector were selected with 2 µg/mL puromycin. Selected cells were expanded and used for ER calcium measurements.

Protein expression and purification

The B1 domain of streptococcal protein G (GB1) and his-tagged versions of wild-type CRT, CRT(1-351), CRT(1-339) and CRTDel52 were cloned into the pGB1 vector (DelProposto et al., 2009). Chemically-competent Origami B(DE3) E. coli cells were transformed with the pGB1 constructs of wild type CRT, CRT(1-351), CRT(1-339) and CRTDel52. Single colonies were grown overnight at 37°C. The overnight cultures were inoculated into larger (1-2 L) volumes of LB media and grown in a shaking incubator at 220 rpm and 37°C until the OD600 of 0.8. Protein expression was then induced with 200 µM IPTG, and the cultures were further incubated at 16°C overnight on shaking. The bacterial cells were harvested, pelleted, and lysed in 50 mL of HEPES buffer containing (20 mM HEPES pH 7.5, 140 mM NaCl, 1 mM CaCl2) by sonication. The lysates were clarified by centrifugation at 10,000 rpm, 4°C for 30 minutes. The supernatant was incubated with nickel-NTA resin (Qiagen, 1018244) for 3 hours at 4°C and purified through a gravity-flow column. The unbound proteins and contaminants were removed by washing with a wash buffer containing 20 mM HEPES pH 7.5, 140 mM NaCl, 1 mM CaCl2, 10 mM imidazole. The his-tagged GB1-CRT proteins were eluted with buffer containing 20 mM HEPES pH 7.5, 140 mM NaCl, 1 mM CaCl2, 350 mM imidazole. The eluted proteins were concentrated and subjected to fast protein liquid chromatography over a Superdex 200 Increase 10/300 GL column at 4°C in 20 mM HEPES pH 7.5, 140 mM NaCl, 1 mM CaCl2 as described previously (Del Cid et al., 2010, Wijeyesakere et al., 2011). Monomer fractions, collected between 14 – 20 mL as shown in Figure 2, were pooled, dialyzed and concentrated to 4 mg/mL in 20 mM HEPES, 10 mM NaCl 1 mM CaCl2 pH 7.5 using Amicon ultra concentrators with a molecular weight cut-off of 30,000 Dalton (Millipore) and frozen in 10% glycerol for further use.

Isothermal titration calorimetry

Isothermal titration calorimetry (ITC) measurements were done using the Nano-ITC SV (TA Instruments). Purified GB1-CRT monomers were dialyzed into 20 mM HEPES, 10 mM NaCl pH 7.5, brought to a concentration of 100 µM, and then pre-incubated with 50-100 µM CaCl2 in 20 mM HEPES, 10 mM NaCl pH 7.5 for 30 minutes to block the high affinity calcium binding sites. This was followed by 25 sequential 10µL injections of 20 mM HEPES, 10 mM NaCl, 10 mM CaCl2 pH 7.5 into the protein solution maintained at 37°C in a total volume of 1.3 mL to final CaCl2 concentrations of 1.6-1.7 mM. The heat changes measured for the sequential injections were curve-fitted to calculate the enthalpies, stoichiometries, and the dissociation constants using NanoAnalyze software.

Cytosolic calcium measurement and store-operated calcium entry (SOCE) induction

MEG-01 cells and their CALR-modified versions were plated on glass coverslips pretreated with 0.01% poly-L-lysine to facilitate cell attachment. In separate experiments, cells were treated with 10 µM Imatinib (Sigma), a tyrosine kinase inhibitor for 2 hours prior to cytosolic calcium measurements. Cells were loaded with 2.5 µM Fura-2/AM dye (Ion Biosciences) supplemented with 0.1% pluronic acid in a physiological salt solution (PSS) containing 140 mM NaCl, 2 mM CaCl2, 5 mM KCl, 2 mM MgCl2, 10 mM HEPES and 11 mM glucose, and incubated for 60 minutes at 37°C. Live single-cell imaging was done at 20X magnification using an Olympus IX71 inverted microscope equipped with a TILL Polychrome V monochromator light source, a Dual-View emission beam splitter, and a cooled Photometrics Quant-EM camera. Cells were transferred to a perfusion chamber that was mounted on the microscope stage, and PSS with or without 2 mM CaCl2 was flowed over cells using a gravity-fed perfusion system. Cytosolic calcium measurements were performed by perfusing Fura-2 loaded cells in the chamber with 2 mM Ca2+, followed by the addition of 25 µM cyclopiazonic acid (CPA) (Cayman) into the imaging buffer. To determine cytosolic calcium levels after activating SOCE, cells were initially perfused with Ca2+-free PSS containing 0.1 mM EGTA, followed by the addition of 25 µM CPA in Ca2+-free PSS and subsequent addition of 2 mM Ca2+ and 25 µM CPA in PSS (without EGTA) to reestablish calcium influx. The emission signals corresponding to the 340 and 380 nm excitation were recorded, and the signals resulting from 340/380 nm excitation were calculated using the Meta Fluor software. The basal cytosolic calcium was defined as the 340/380 nm signal ratio before CPA treatment (measured over a 4- or 5-minute interval immediately before CPA addition). The fractional increase in cytosolic calcium levels after ER Ca2+ depletion with CPA was calculated as the difference between peak signal (single time point) and baseline (averaged over a 4- or 5-minute interval before CPA addition), as a ratio relative to the baseline. The fractional increase in cytosolic calcium after addition of 2 mM extracellular Ca2+ and CPA was calculated as the difference between the peak signal following Ca2+ and CPA addition (averaged over a 1-minute time interval following the peak signal) and the baseline (the signal averaged over a 0.8-1-minute interval immediately before the Ca2+ and CPA addition) relative to the baseline. Total calcium release after ER Ca2+ depletion was reported as the total area under the curve (AUC). Fractional increases were calculated as (Peak-Baseline)/Baseline × 100. This was calculated as peak signals after CPA addition relative to the baseline before CPA (measured over a 4- or 5-minute interval immediately before CPA addition).

ER calcium measurements in HEK293T cells

CALR-modified HEK cells as described in methods were transiently transfected with pCIS-GEM-CEPIA1er (Suzuki et al., 2014); Addgene plasmid # 58217) using Fugene HD reagent (Promega). The transfected cells were harvested after 72 hours. Cells were washed twice with FACS buffer (PBS + 2% FBS) supplemented with 2 mM Ca2+ and resuspended in the same FACS buffer. The live-dead stain 7-AAD (BD) was added at 1:200 dilution and the cells were analyzed by flow cytometry using the BD LSRFortessa cell analyzer. The signal for the Ca2+-bound GEM-CEPIA1er probe was recorded in the Pacific Blue channel (excitation at 405 nm and emission at 452-455 nm) and the signal for the unbound GEM-CEPIA1er probe was recorded in the AmCyan channel (excitation at 405 and emission at 498 nm). The analysis of flow cytometry data was done in the FlowJo software (v10.10.00). The cells were gated based on the forward (FSC-A) and side scatter (SSC-A), followed by gating on the 7AAD-negative (live) cells, and subsequent gating on the cells positive for the expression of GEM-CEPIA1er probe measured in the Pacific Blue and AmCyan channels. Basal signals were initially recorded, followed by the addition of 5 µM thapsigargin (Tg) and subsequent recording of signals for the treated and untreated cells at four different time points between 2 – 10 minutes. The ratio of mean fluorescence intensity (MFI) of the bound to the unbound signal was used for calculations. In additional experiments, ER calcium levels were measured by spectrofluorimetry in HEK cells transfected with pCIS-GEM-CEPIA1er. Fluorescence emission spectra were recorded between 420-600 nm, following excitation at 390 nm. Basal spectra were initially recorded using 3 million cells suspended in phosphate-buffered saline (PBS) containing 2 mM Ca2+, followed by treatment with 5 µM Tg, and additional spectra were recorded at 3 min intervals (3 - 15) mins. The ratios of mean fluorescence intensities of the bound (∼ 460 nm) to unbound (∼ 510 nm) peak signals were calculated in the absence of Tg or at 15 mins following Tg addition to compare the basal signals and the changes following Tg addition.

Immunoblotting

The HEK293T and MEG-01 cells were harvested and lysed in lysis buffer (50 mM Tris pH 7.5, 150 mM NaCl, 1% Triton X-100, 5 mM CaCl2) for 30 minutes at 4°C. The lysates were clarified by centrifugation, and the total amount of protein in the lysates was determined using the BCA reagent kit (Thermoscientific). The volume of lysate containing 20-40 μg of protein was denatured by boiling in 1X SDS loading dye for 5 minutes and subjected to SDS-PAGE. Following electrophoresis, the proteins were transferred onto a PVDF membrane. The membranes were blocked in 5% non-fat dry milk solution in TBST (Tris-buffered saline with Tween-20), followed by immunoblotting and detection for protein expression using primary anti-CRT(N) antibody (Cell Signaling Technology (CST) 12238, 1:10,000 dilution) for detection of CRTWT and anti-CRT(C-mut) antibody for detection of CRT mutants (Venkatesan et al., 2021a). Primary anti-STAT5 (94205S, 1:1000), anti-phospho-STAT5 (93559S, 1:1000), anti-GRP78/BiP (11587-1-AP, 1:10,000), anti-GRP94 (14700-1-AP, 1:10,000), anti-ERp57 (15967-1-AP, 1:10,000) and anti-PDIA6 (18233-1-AP, 1:20,000) were used to probe for the expression of BiP, GRP94, PDIA3 and PDIA6 respectively. Anti-GAPDH (CST 2118, 1:20,000) and anti-vinculin (CST 13901, 1:20,000) were used to probe for GAPDH and vinculin as a loading control. Blots were incubated with horseradish peroxidase-conjugated secondary antibodies (Jackson ImmunoResearch) and developed by chemiluminescence (Pierce™ ECL western blotting substrate, 32106). Immunoblot images were captured using the Azure 300 chemi imager (Azure Biosystems).

RNA-sequencing

The parental and CRT-KO MEG-01 cells were grown in complete media for 12-18 hours before harvesting. Total RNA was isolated using the RNeasy Plus Mini Kit (Qiagen). This was followed by poly-A selection and enrichment of mRNA using NEB Next Ultra Express RNA kit and a NEB Poly A kit (New England Biolabs). Sequencing was done on the NovaSeq at PE150, targeting 30-40 million reads/sample after cDNA library preparation. Data were pre-filtered to remove genes with 0 counts in all samples. Differential gene expression analysis was performed using DESeq2. The differentially expressed genes were selected as those with a log2 fold-change of <-0.49 or >0.49 and p values <0.05. The results from RNA sequencing were further analyzed in the iPathwayGuide to identify significantly altered proteins related to calcium signaling. The volcano plot was plotted in GraphPad Prism as Log2 fold-change (x-axis) vs. -Log10 adjusted p-value.

Real-time polymerase chain reaction

Total RNA was isolated using the RNeasy Plus Mini Kit (Qiagen) followed by synthesis of complementary DNA using the Superscript III First-Strand Synthesis Kit (ThermoFisher Scientific) following the manufacturer’s protocol. Quantitative RT-PCR was performed on the ABI 7500 FAST Real Time PCR machine (ThermoFisher Scientific, USA) using the SYBR Green PCR Master mix (Applied Biosystems, UK) and gene-specific primers (Supplementary Table 2). Measurement was done in technical duplicates. The expression of target genes was quantified using the 2−ΔΔCt method and normalized to the expression of either one or multiple endogenous control genes including beta-actin (ACTB), glyceraldehyde 3-phosphate dehydrogenase (GAPDH) and hypoxanthine phosphoribosyltransferase 1 (HPRT1). The expression of the target gene in MEG-01 CRT-KO and wild type and CRTDel52 reconstitution cells was further normalized to the mean expression of parental MEG-01 or CRT-KO-vec. The calculated relative mRNA transcript expression values (RQ) were plotted using GraphPad Prism.

Statistical analysis

All data were compiled using Microsoft Excel and analyzed using GraphPad Prism 10.4. A Student’s t-test was used to compare mean differences between two groups, while a one-way ANOVA was used to compare mean differences among more than two groups. The threshold for statistical significance was observed at a p-value ≤ 0.05.